# Connecting Docker, PostgreSQL, and R {#chapter_connect-docker-postgresql-r}

> This chapter demonstrates how to:
>
>  * Run, clean-up and close PostgreSQL in Docker containers.
>  * Keep necessary credentials secret while being available to R when it executes.
>  * Interact with PostgreSQL when it's running inside a Docker container.
>  * Read and write to PostgreSQL from R.

Please install the `sqlpetr` package if not already installed:
```{r sqlpetr, echo = TRUE, message=FALSE, warning=FALSE}
library(devtools)
if (!require(sqlpetr)) devtools::install_github("smithjd/sqlpetr")
```
Note that when you install the package the first time, it will ask you to update the packages it uses and that can take some time.

The following packages are used in this chapter:
```{r setup, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(DBI)
library(RPostgres)
require(knitr)
library(sqlpetr)
```

## Verify that Docker is running

Docker commands can be run from a terminal (e.g., the Rstudio Terminal pane) or with a `system2()` command.  (We discuss the diffeent ways of interacting with Docker and other elements in your environment in a [separate chapter](#your-local-environment).)  The necessary functions to start, stop Docker containers and do other busy work are provided in the `sqlpetr` package.  As time permits and curiosity dictates, feel free to look at those functions to see how they work.

Check that Docker is up and running:

```{r}
sp_check_that_docker_is_up()
```

## Remove previous containers if they exist
Remove the `cattle` and `sql-pet` containers if they exist (e.g., from prior experiments).  
```{r docker cleanup}
sp_docker_remove_container("cattle")
sp_docker_remove_container("sql-pet")
```

We name containers `cattle` for "throw-aways" and `pet` for ones we treasure and keep around.  :-)
```{r setup cattle Docker container, eval=TRUE, echo=TRUE}
sp_make_simple_pg("cattle")

```

Docker returns a long string of numbers.  If you are running this command for the first time, Docker downloads the PostgreSQL image, which takes a bit of time.

The following command shows that a container named `cattle` is running `postgres:10`.  `postgres` is waiting for a connection:
```{r }
sp_check_that_docker_is_up()

```
## Connect, read and write to PostgreSQL from R


### Connect with PostgreSQL

Connect to PostgreSQL using the `sp_get_postgres_connection` function:
```{r}
con <- sp_get_postgres_connection(user = "postgres",
                         password = "postgres",
                         dbname = "postgres",
                         seconds_to_test = 30)
```
Notice that we are using the PostgreSQL default username and password at this point and that it's in plain text. That is bad practice because user credentials should not be shared in open code like that.  A [subsequent chapter](#dbms-login) demonstrates how to store and use credentials to access the dbms so that they are kept private.

Make sure that you can connect to the PostgreSQL database that you have just started. If you have been executing the code from this tutorial, the database will not contain any tables yet:

```{r}
DBI::dbListTables(con)
```

### Interact with PostgreSQL

Write `mtcars` to PostgreSQL
```{r}
DBI::dbWriteTable(con, "mtcars", mtcars, overwrite = TRUE)
```

List the tables in the PostgreSQL database to show that `mtcars` is now there:

```{r}
DBI::dbListTables(con)
```

List the fields in mtcars:
```{r}
DBI::dbListFields(con, "mtcars")
```

Download the table from the DBMS to a local data frame:
```{r}
mtcars_df <- DBI::dbReadTable(con, "mtcars")
```

Show a few rows:
```{r}
sp_print_df(head(mtcars_df))

```

## Clean up

Afterwards, always disconnect from the dbms:
```{r}
DBI::dbDisconnect(con)
```
Tell Docker to stop the `cattle` container:
```{r}
sp_docker_stop("cattle")
```

Tell Docker to *remove* the `cattle` container from it's library of active containers:
```{r}
sp_docker_remove_container("cattle")
```

If we just **stop** the Docker container but don't remove it (as we did with the `sp_docker_remove_container("cattle")` command), the `cattle` container will persist and we can start it up again later with `sp_docker_start("cattle")`.  In that case, `mtcars` would still be there and we could retrieve it from PostgreSQL again.  Since `sp_docker_remove_container("cattle")`  has removed it, the updated database has been deleted.  (There are enough copies of `mtcars` in the world, so no great loss.)
