# Create the dvdrental database in PostgreSQL in Docker {#chapter_setup-dvdrental-db}

> This chapter demonstrates how to:
>
>  * Setup the `dvdrental` database in Docker
>  * Stop and start Docker container to demonstrate persistence
>  * Connect to and disconnect R from the `dvdrental` database
>  * Set up the environment for subsequent chapters

## Overview

In the last chapter we connected to PostgreSQL from R.  Now we set up a "realistic" database named `dvdrental`. There are different approaches to doing this: this chapter sets it up in a way that doesn't show all the Docker details.

These packages are called in this Chapter:
```{r setup, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(DBI)
library(RPostgres)
library(glue)
require(knitr)
library(dbplyr)
library(sqlpetr)
library(bookdown)
```

## Verify that Docker is up and running
```{r docker verify}
sp_check_that_docker_is_up()
```

## Clean up if appropriate
Force-remove the `cattle` and `sql-pet` containers if they exist (e.g., from a prior runs):
```{r}
sp_docker_remove_container("cattle")
sp_docker_remove_container("sql-pet")
```
## Build the pet-sql Docker image
For the rest of the book we will be using a Docker image called
`postgres-dvdrental`. To save space here in the book, we've created a function
in `sqlpetr` to build this image, called [`sp_make_dvdrental_image`](https://smithjd.github.io/sqlpetr/reference/sp_make_dvdrental_image.html). Vignette [Building the `dvdrental` Docker Image
](https://smithjd.github.io/sqlpetr/articles/building-the-dvdrental-docker-image.html) describes the build process.

```{r}
sp_make_dvdrental_image("postgres-dvdrental")
```

Did it work? We have a function that lists the images into a tibble!
```{r}
sp_docker_images_tibble()

```

## Run the pet-sql Docker Image
Now we can run the image in a container and connect to the database. To run the
image we use an `sqlpetr` function called [`sp_pg_docker_run`](https://smithjd.github.io/sqlpetr/reference/sp_pg_docker_run.html)

When the image runs in the container, we can mount the current working directory
into a path in the container. You'll see this in action in a future chapter. 
Docker will create this path if it doesn't exist.

To specify the path, set the parameter `mount_here_as` to the name you want.
Rules for the name:

* If you don't want to mount into the container, specify `NULL`. This is the 
default!
* The name must start with a `/` and be a valid absolute path.
The name should contain only slashes, letters, numbers and underscores. Other characters may or may not work. The `snakecase` package is your friend.

```{r}
sp_pg_docker_run(
  container_name = "sql-pet",
  image_tag = "postgres-dvdrental",
  postgres_password = "postgres",
  mount_here_as = "/petdir"
)
```

Did it work?
```{r}
sp_docker_containers_tibble()

```

## Connect to PostgreSQL with R

Use the DBI package to connect to the `dvdrental` database in PostgreSQL.  Remember the settings discussion about [keeping passwords hidden][Pause for some security considerations]

```{r }
con <- sp_get_postgres_connection(
  host = "localhost",
  port = 5432,
  user = "postgres",
  password = "postgres",
  dbname = "dvdrental",
  seconds_to_test = 30
)
```

List the tables in the database and the fields in one of those tables.  
```{r }
dbListTables(con)
```
```{r }
dbListFields(con, "rental")
```

Disconnect from the database:
```{r }
dbDisconnect(con)

```
## Stop and start to demonstrate persistence

Stop the container:
```{r}
sp_docker_stop("sql-pet")
sp_docker_containers_tibble()
```

When we stopped `sql-pet`, it no longer appeared in the tibble. But the
container is still there. `sp_docker_containers_tibble` by default only lists
the *running* containers. But we can use the `list_all` option and see it:

```{r}
sp_docker_containers_tibble(list_all = TRUE)
```


Restart the container and verify that the dvdrental tables are still there:
```{r}
sp_docker_start("sql-pet")
sp_docker_containers_tibble()
```
Connect to the `dvdrental` database in PostgreSQL:
```{r}
con <- sp_get_postgres_connection(
  host = "localhost",
  port = 5432,
  user = "postgres",
  password = "postgres",
  dbname = "dvdrental",
  seconds_to_test = 30
)
```

Check that you can still see the fields in the `rental` table:
```{r}
dbListFields(con, "rental")
```

## Cleaning up

Always have R disconnect from the database when you're done.
```{r}

dbDisconnect(con)

```

Stop the `sql-pet` container:
```{r}
sp_docker_stop("sql-pet")
```
Show that the container still exists even though it's not running

```{r}
sp_show_all_docker_containers()

```

Next time, you can just use this command to start the container: 

> `sp_docker_start("sql-pet")`

And once stopped, the container can be removed with:

> `sp_check_that_docker_is_up("sql-pet")`

## Using the `sql-pet` container in the rest of the book

After this point in the book, we assume that Docker is up and that we can always start up our *sql-pet database* with:

> `sp_docker_start("sql-pet")`
