# Asking Business Questions From a Single Table {#chapter_exploring-a-single-table}

> This chapter explores:
>
>   * Issues that come up when investigating a single table from a business perspective
>   * Show the multiple data anomalies found in a single AdventureWorks table (*salesorderheader*)
>   * The interplay between "data questions" and "business questions"

The previous chapter has demonstrated some of the automated techniques for showing what's in a table using some standard R functions and packages.  Now we demonstrate a step-by-step process of making sense of what's in one table with more of a business perspective.  We illustrate the kind of detective work that's often involved as we investigate the *organizational meaning* of the data in a table.  We'll investigate the `salesorderheader` table in the `sales` schema with a key business question in mind; are sales dollars trending up, down or flat?

For this kind of detective work we are seeking to understand the following elements separately and as they interact with each other:

  * What data is stored in the database and 
  * How information is represented
  * How the data is entered at a day-to-day level to represent business activities
  * How the business itself is changing over time

## Setup our standard working environment

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 10)
sleep_default <- 3
```

Use these libraries:
```{r libraries, message=FALSE}
library(tidyverse)
library(DBI)
library(RPostgres)
library(connections)
library(glue)
require(knitr)
library(dbplyr)
library(sqlpetr)
library(bookdown)
library(here)
library(lubridate)
library(gt)
library(scales)
theme_set(theme_light())
```

Connect to `adventureworks`.  In an interactive session we prefer to use `connections::connection_open` instead of dbConnect


```{r, start adventureworks and connect}
#sp_docker_start("adventureworks")
Sys.sleep(sleep_default)
# con <- dbConnect(
con <- connection_open(
  RPostgres::Postgres(),
  bigint = "integer",  
  host = "localhost",
  port = 5432,
  user = "postgres",
  password = "postgres",
  dbname = "adventureworks")
```

Some queries generate big integers, so we need to include `RPostgres::Postgres()` and `bigint = "integer"` in the connections statement.


## A word on naming 

> You will find that many tables will have columns with the same name in an enterprise database.  For example, in the adventureworks database, almost all tables have columns named `rowguid` and `modifieddate` and there are many other examples of names that are reused throughout the database.  The meaning of a column depends on the table that contains it, so as you pull a column out of a table, naming its provenance is important.
>
> Naming columns carefully (whether retrieved from the database or calculated)  will pay off, especially as our queries become more complex. Using `soh` as an abbreviation of *sales order header* to tag statistics that are derived from the `salesorderheader` table, as we do in this book, is one example of an intentional naming strategy: it reminds us of the original source of a column.  You, future you, and your collaborators will appreciate the effort although different naming conventions are completely valid.  And a naming convention when rigidly applied can yield some long and ugly names.
>
> In the following example `soh` appears in different positions in the column name but it is easy to guess at a glance that the data comes from the `salesorderheader` table.
>
> Naming derived tables is just as important as naming derived columns.

## The overall AdventureWorks sales picture

We begin by looking at Sales on a yearly basis, then consider monthly sales.  We discover that half way through the period represented in the database, the business appears to begin selling online, which has very different characteristics than sales by Sales Reps.  We then look at the details of how Sales Rep sales are recorded in the system and discover a data anomaly that we can correct. 

## Annual sales

On an annual basis, are sales dollars trending up, down or flat? We begin with annual revenue and number of orders.  

```{r Calculate time period and annual sales dollars 1}
annual_sales <- tbl(con, in_schema("sales", "salesorderheader")) %>%
  mutate(year = substr(as.character(orderdate), 1, 4)) %>%
  group_by(year) %>%
  summarize(
    min_soh_orderdate = min(orderdate, na.rm = TRUE),
    max_soh_orderdate = max(orderdate, na.rm = TRUE),
    total_soh_dollars = round(sum(subtotal, na.rm = TRUE), 2),
    avg_total_soh_dollars = round(mean(subtotal, na.rm = TRUE), 2),
    soh_count = n()
  ) %>%
  arrange(year) %>%
  select(
    year, min_soh_orderdate, max_soh_orderdate, total_soh_dollars,
    avg_total_soh_dollars, soh_count
  ) %>% 
  show_query() %>% 
  collect() 
  
```

Note that all of this query is running on the server.  The `show_query` function shows the SQL code postgreSQL is runs to produce its output.

```{r Show the column format of the resulting query}
annual_sales %>% str()
```

We hang on to some date information for later use in plot titles.

```{r You will need time periods later on.}
min_soh_dt <- min(annual_sales$min_soh_orderdate)
max_soh_dt <- max(annual_sales$max_soh_orderdate)
```

### Total sales by year

```{r Calculate time period and annual sales dollars - 2 , fig.height=4, fig.width=4}
ggplot(data = annual_sales, aes(x = year, y = total_soh_dollars)) +
  geom_col() +
  geom_text(aes(label = round(as.numeric(total_soh_dollars), digits = 0)), vjust = -0.25) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Adventure Works Sales Dollars by Year",
    x = glue("Year - between ", {format(min_soh_dt, "%B %d, %Y")} , " and  ", 
            {format(max_soh_dt, "%B %d, %Y")}),
    y = "Sales $"
  )
```
Both 2011 and 2014 are shorter time spans than the other two years, making comparison interpretation more difficult. 

Both 2011 and 2014 are shorter time spans than the other two years, making comparison across the years more difficult. Could it be that sales are seasonal? Let's look at the order volume and see whether there's a pattern in the sales data.

### Total order volume

Look at number of orders per year:
```{r average dollars per sale - v2, fig.height=4, fig.width=4}
ggplot(data = annual_sales, aes(x = year, y = as.numeric(soh_count))) +
  geom_col() +
  geom_text(aes(label = round(as.numeric(soh_count), digits = 0)), vjust = -0.25) +
  labs(
    title = "Number of orders per year",
    x = glue("Years between ", {format(min_soh_dt, "%B %d, %Y")} , " and  ", 
            {format(max_soh_dt, "%B %d, %Y")}),
    y = "Total Number of Orders"
  )
```

That's a huge jump in the number of orders between 2012 and 2013.  Given the total annual dollars, we ask whether the size of a sale has changed.

### Average dollars per sale
```{r average dollars per sale - , fig.height=4, fig.width=4}
ggplot(data = annual_sales, aes(x = year, y = avg_total_soh_dollars)) +
  geom_col() +
  scale_y_continuous(labels = scales::dollar_format()) +
  geom_text(aes(label = round(avg_total_soh_dollars, digits = 0)), vjust = -0.25) +
  labs(
    title = "Average Dollars per Sale",
    x = glue("Year - between ", {format(min_soh_dt, "%B %d, %Y")} , " to  ", 
            {format(max_soh_dt, "%B %d, %Y")}),
    y = "Average Sale Amount"
  )
```

That's a big drop between average sale of more than $7,000 to less than $3,000.  A remarkable change has taken place in this business.


From 2012 to 2013 the average dollars per order dropped from more than $8,500 to nearly $3,000 while the total number of orders shot up from less than 4,000 to more than 14,000.  **Why are the number of orders increasing, but the average dollar amount of a sale is dropping?  **

We need to drill down to look at monthly sales, adapting the first query to group by month and year.

## Monthly Sales

Our next iteration drills down from annual sales dollars to monthly sales dollars. For that we download the orderdate as a date, rather than a character variable for the year.  R handles the conversion from the PostgreSQL date-time to an R date-time.  We then convert it to a simple date with a `lubridate` function.

The following query uses the [postgreSQL function `date_trunc`](https://www.postgresqltutorial.com/postgresql-date_trunc/), which is equivalent to `lubridate`'s `round_date` function in R.  If you want to push as much of the processing as possible onto the database server and thus possibly deal with smaller datasets in R, interleaving [postgreSQL functions](https://www.postgresql.org/docs/current/functions.html) into your dplyr code will help.

```{r monthly sales retrieval - sql}
monthly_sales <- tbl(con, in_schema("sales", "salesorderheader")) %>%
  select(orderdate, subtotal) %>%
  mutate(
    orderdate = date_trunc('month', orderdate)
  ) %>%
  group_by(orderdate) %>%
  summarize(
    total_soh_dollars = round(sum(subtotal, na.rm = TRUE), 2),
    avg_total_soh_dollars = round(mean(subtotal, na.rm = TRUE), 2),
    soh_count = n()
  ) %>% 
  collect() 
```

In many cases we don't really care where our queries are executed, but if we do we have to look for cases where functions from R packages like `lubridate` and the equivalent `postgreSQL` functions are subtly different.  In this case the `postgreSQL` function produces a `POSIXct` column, not a `Date` so we need to tack on a mutate function onto our query.

>
> When to use the sql() function.

```{r}
monthly_sales <-  monthly_sales %>% 
  mutate(orderdate = as.Date(orderdate))
```

Plotting the monthly sales data:

```{r Total monthly sales bar chart}

ggplot(data = monthly_sales, aes(x = orderdate, y = total_soh_dollars)) +
  geom_col() +
  scale_y_continuous(labels = dollar) +
  theme(plot.title = element_text(hjust = 0.5)) + # Center the title
  labs(
    title = glue("Sales by Month\n", {format(min_soh_dt, "%B %d, %Y")} , " to  ", 
            {format(max_soh_dt, "%B %d, %Y")}),
    x = "Month",
    y = "Sales Dollars"
  )
```

### Check lagged monthly data

The total sales are trending up but suspiciously uneven.  We'll use `dplyr::lag` to help with our month over month compairson and will later visualize just how much month-to-month difference there is:

```{r}
monthly_sales_lagged <- monthly_sales %>%
  mutate(monthly_sales_change = (dplyr::lag(total_soh_dollars)) - total_soh_dollars)
```

```{r}
(sum_lags <- summary(monthly_sales_lagged$monthly_sales_change))
```

The trend is positive on average `r prettyNum(sum_lags[["Mean"]], big.mark = ",", format = "f", digits = 2)` but half of the months have swings greater than `r prettyNum(sum_lags[["3rd Qu."]] - sum_lags[["1st Qu."]], big.mark = ",", format = "f", digits = 2)`!


```{r, message=FALSE, error=FALSE}
ggplot(monthly_sales_lagged, aes(x = orderdate, y = monthly_sales_change)) +
  scale_x_date(date_breaks = "year", date_labels = "%Y", date_minor_breaks = "3 months") +
  geom_line() +
  geom_point() +
  scale_y_continuous( labels = scales::dollar_format()) +
  # scale_y_continuous(limits = c(-1700000,4500000), labels = scales::dollar_format()) +
  theme(plot.title = element_text(hjust = .5)) + 
  labs(
    title = glue(
      "Monthly Sales Change \n",
      "Between ", {format(min_soh_dt, "%B %d, %Y")} , " and  ", 
            {format(max_soh_dt, "%B %d, %Y")}
    ),
    x = "Month",
    y = "Dollar Change"
  )
```

AdventureWorks sales are *very* uneven.  We'll come back to this issue shortly.

### Comparing dollars and orders to a base year

To look at dollars and the number of orders together, we compare the monthly data to the yearly average for 2011.

```{r}
start_year <- monthly_sales %>%
  mutate(yr = year(orderdate)) %>%
  group_by(yr) %>%
  summarize(
    total_soh_dollars = sum(total_soh_dollars),
    soh_count = sum(soh_count),
    n_months = n(),
    avg_dollars = total_soh_dollars / n_months,
    avg_count = soh_count / n_months
  ) %>%
  filter(yr == min(yr))
```

Use 2011 as a baseline:

```{r}
start_year
```

Re express monthly data in terms of the baseline and plot:

```{r}
monthly_sales_base_year_normalized_to_2011 <- monthly_sales %>%
  mutate(
    dollars = (100 * total_soh_dollars) / start_year$avg_dollars,
    number_of_orders = (100 * soh_count) / start_year$avg_count
  ) %>%
  ungroup()

monthly_sales_base_year_normalized_to_2011 <- monthly_sales_base_year_normalized_to_2011 %>%
  select(orderdate, dollars, `# of orders` = number_of_orders) %>%
  pivot_longer(-orderdate,
    names_to = "relative_to_2011_average",
    values_to = "amount"
  )

monthly_sales_base_year_normalized_to_2011 %>%
  ggplot(aes(orderdate, amount, color = relative_to_2011_average)) +
  geom_line() +
  geom_hline(yintercept = 100) +
  scale_x_date(date_labels = "%Y-%m", date_breaks = "6 months") +
  labs(
    title = glue(
      "Adventureworks Normalized Monthly Sales\n",
      "Number of Sales Orders and Dollar Totals\n",
      {format(min_soh_dt, "%B %d, %Y")} , " to  ", 
            {format(max_soh_dt, "%B %d, %Y")}),
    x = "Date",
    y = "",
    color = "% change from\n 2011 average"
  ) +
  theme(legend.position = c(.3,.75))
```

## The effect of online sales

We suspect that the business has changed a lot with the advent of online orders so we check the impact of `onlineorderflag` on annual sales.  The `onlineorderflag` indicates which sales channel accounted for the sale, **Sales Reps** or **Online**.

### Add `onlineorderflag` to our annual sales query

```{r Calculate time period and annual sales dollars- 3}
annual_sales_w_channel <- tbl(con, in_schema("sales", "salesorderheader")) %>%
  select(orderdate, subtotal, onlineorderflag) %>%
  collect() %>%
  mutate(
    orderdate = date(orderdate),
    orderdate = round_date(orderdate, "year"),
    onlineorderflag = if_else(onlineorderflag == FALSE,
      "Sales Rep", "Online"
    ),
    onlineorderflag = as.factor(onlineorderflag)
  ) %>%
  group_by(orderdate, onlineorderflag) %>%
  summarize(
    min_soh_orderdate = min(orderdate, na.rm = TRUE),
    max_soh_orderdate = max(orderdate, na.rm = TRUE),
    total_soh_dollars = round(sum(subtotal, na.rm = TRUE), 2),
    avg_total_soh_dollars = round(mean(subtotal, na.rm = TRUE), 2),
    soh_count = n()
  ) %>%
  select(
    orderdate, onlineorderflag, min_soh_orderdate,
    max_soh_orderdate, total_soh_dollars,
    avg_total_soh_dollars, soh_count
  )
```

### Annual Sales comparison

Start by looking at total sales.

```{r Calculate annual sales dollars}

ggplot(data = annual_sales_w_channel, aes(x = orderdate, y = total_soh_dollars)) +
  geom_col() +
  scale_y_continuous(labels = scales::dollar_format()) +
  facet_wrap("onlineorderflag") +
  labs(
    title = "Adventure Works Sales Dollars by Year",
    caption = glue( "Between ", {format(min_soh_dt, "%B %d, %Y")} , " - ", 
            {format(max_soh_dt, "%B %d, %Y")}),
    subtitle = "Comparing Online and Sales Rep sales channels",
    x = "Year",
    y = "Sales $"
  )
```

Indeed the total sales are quite different as are the number of orders and the average order size!

### Order volume comparison

Look at number of orders per year:
```{r average dollars per sale - v4}
ggplot(data = annual_sales_w_channel, aes(x = orderdate, y = as.numeric(soh_count))) +
  geom_col() +
  facet_wrap("onlineorderflag") +
  labs(
    title = "Adventure Works Number of orders per Year",
    caption = glue( "Between ", {format(min_soh_dt, "%B %d, %Y")} , " - ", 
            {format(max_soh_dt, "%B %d, %Y")}),
    subtitle = "Comparing Online and Sales Rep sales channels",
    x = "Year",
    y = "Total number of orders"
  )
```

### Comparing **Sales Rep** sales to **Online** Orders

```{r average dollars per sale 3}
ggplot(data = annual_sales_w_channel, aes(x = orderdate, y = avg_total_soh_dollars)) +
  geom_col() +
  facet_wrap("onlineorderflag") +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Average Dollars per Sale",
    x = glue( "Year - between ", {format(min_soh_dt, "%B %d, %Y")} , " - ", 
            {format(max_soh_dt, "%B %d, %Y")}),
    y = "Average sale amount"
  )
```


## Impact of order type on monthly sales

Digging into the difference between **Sales Rep** and **Online** sales. 

### Retrieve monthly sales with the `onlineorderflag` 

This query puts the `collect` statement earlier than the previous queries.

```{r}
monthly_sales_w_channel <- tbl(con, in_schema("sales", "salesorderheader")) %>%
  select(orderdate, subtotal, onlineorderflag) %>%
  collect() %>% # From here on we're in R

  mutate(
    orderdate = date(orderdate),
    orderdate_rounded = round_date(orderdate, "month"),
    onlineorderflag = if_else(onlineorderflag == FALSE,
      "Sales Rep", "Online"
    ),
  ) %>% #
  group_by(orderdate, onlineorderflag) %>%
  summarize(
    min_soh_orderdate = min(orderdate, na.rm = TRUE),
    max_soh_orderdate = max(orderdate, na.rm = TRUE),
    total_soh_dollars = round(sum(subtotal, na.rm = TRUE), 2),
    avg_total_soh_dollars = round(mean(subtotal, na.rm = TRUE), 2),
    soh_count = n()
  ) %>%
  ungroup()
```


```{r}
monthly_sales_w_channel %>%
  rename(`Sales Channel` = onlineorderflag) %>%
  group_by(`Sales Channel`) %>%
  summarize(
    unique_dates = n(),
    start_date = min(min_soh_orderdate),
    end_date = max(max_soh_orderdate),
    total_sales = sum(total_soh_dollars)
  ) %>%
  gt()
```

As we will see later on, the **Sales Rep** data doesn't match the **Online** data.  The **Online** data includes 2 months that are not included in the main sales channel.

### Monthly variation compared to a trend line

Jumping to the trend line comparison, we see the source of the variation.

```{r average dollars, warning=FALSE}

ggplot(
  data = monthly_sales_w_channel,
  aes(
    x = orderdate, y = total_soh_dollars
  )
) +
  geom_line() +
  geom_smooth(se = FALSE) +
  facet_grid("onlineorderflag", scales = "free") +
  scale_y_continuous(labels = dollar) +
  scale_x_date(date_breaks = "year", date_labels = "%Y", date_minor_breaks = "3 months") +
  theme(plot.title = element_text(hjust = .5)) + # Center ggplot title
  labs(
    title = glue(
      "Sales by Month by Year"
    ),
    x = glue( "Month - between ", {format(min_soh_dt, "%B %d, %Y")} , " - ", 
           {format(max_soh_dt, "%B %d, %Y")}),
    y = "Sales Dollars"
  )
```

The **monthly** gyrations are happening on the Sales Rep side, amounting to differences in a million dollars compared to small variations of around $25,000.

### Compare monthly lagged data by order type

First consider month-to-month change.
```{r , warning=FALSE}
monthly_sales_w_channel_lagged_by_month <- monthly_sales_w_channel %>%
  ungroup() %>% 
  group_by(onlineorderflag) %>%
  mutate(
    pct_monthly_soh_dollar_change =
      total_soh_dollars / (lag(total_soh_dollars, 1)) * 100,
    pct_monthly_soh_count_change =
      soh_count / (lag(soh_count, 1)) * 100
  )

ggplot(monthly_sales_w_channel_lagged_by_month, aes(x = orderdate, y = pct_monthly_soh_dollar_change)) +
  scale_x_date(date_breaks = "year", date_labels = "%Y", date_minor_breaks = "3 months") +
  scale_y_continuous() +
  facet_grid("onlineorderflag", scale = "free") +
  geom_line() +
  theme(plot.title = element_text(hjust = .5)) + # Center ggplot title
  labs(
    title = glue(
      "Monthly Percent Sales Change \n",
      "Comparing Online to Sales Rep Sales"
    ),
    x = paste0("Month - between ", 
               glue({format(min_soh_dt, "%B %d, %Y")} , " - ", 
           {format(max_soh_dt, "%B %d, %Y")})),
    y = "% Dollar Change"
  )

summary(monthly_sales_w_channel_lagged_by_month)
```

For **Sales Reps** it looks like the variation is in the number of orders, not just dollars, as shown in the following plot.
```{r , warning=FALSE}
ggplot(monthly_sales_w_channel_lagged_by_month, aes(x = orderdate, y = pct_monthly_soh_count_change)) +
  scale_x_date(date_breaks = "year", date_labels = "%Y", date_minor_breaks = "3 months") +
  facet_grid("onlineorderflag" , scales = "free") +
  geom_line() +
  theme(plot.title = element_text(hjust = .5)) + # Center ggplot title
  labs(
    title = glue(
      "Monthly Order Volume Change \n",
      "Comparing Online to Sales Rep Sales\n",
      glue({format(min_soh_dt, "%B %d, %Y")} , " - ", 
           {format(max_soh_dt, "%B %d, %Y")})
    ),
    x = "Month",
    y = "Change number of orders"
  )
```

Let's examine whether there is a large year-to-year change.

```{r , warning=FALSE}
monthly_sales_w_channel_lagged_by_year <- monthly_sales_w_channel %>%
  group_by(onlineorderflag) %>%
  mutate(
    pct_monthly_soh_dollar_change =
      total_soh_dollars / (lag(total_soh_dollars, 12)) * 100,
    pct_monthly_soh_count_change =
      soh_count / (lag(soh_count, 12)) * 100
  )

ggplot(
  monthly_sales_w_channel_lagged_by_year,
  aes(x = orderdate, y = pct_monthly_soh_dollar_change)
) +
  scale_x_date(
    date_breaks = "year", date_labels = "%Y",
    date_minor_breaks = "3 months"
  ) +
  scale_y_continuous(limits = c(-10, 300)) +
  facet_grid("onlineorderflag") +
  geom_line() +
  theme(plot.title = element_text(hjust = .5)) + # Center ggplot title
  labs(
    title = glue(
      "Year-on-Year Total Monthly Sales Change \n",
      "Comparing Online to Sales Rep Sales"
    ),
    x = paste0("Month - between ", min_soh_dt, " - ", max_soh_dt),
    y = "% Dollar Change"
  )
```

That's much smaller than the month-to-month change.

> ??? Comparing the number of sales orders year over year by month for 2013 and 2012, one can see that the 2013 sales are between 1.2 and 1.8 times larger than the corresponding month of 2012 from January through June.  In July the 2013 sales are 5 to 6 times the 2012 sales orders. ???

This trend continues into 2014 before the number of sales plummet to just 1.3 time in June.

We suspect that the business has changed a lot with the advent of **Online** orders.

## Detect and diagnose the day of the month problem

Looking at the raw data leads us to suspect that there is a problem with the dates on which Sales Rep orders are entered.

### Sales Rep Orderdate Distribution

Look at the dates when sales are entered for sales by **Sales Reps**.  The following query merits some discussion.

Digging into the day-of-month problem.

```{r}
sales_rep_day_of_month_sales <- tbl(con, in_schema("sales", "salesorderheader")) %>%
  filter(onlineorderflag == FALSE) %>% # Drop online orders
  select(orderdate, subtotal) %>%
  mutate(
    year = year(orderdate),
    month = month(orderdate),
    day = day(orderdate)
  ) %>%
  count(year, month, day) %>% 
  arrange(year, month) %>% 
  collect() %>% 
  ungroup() %>% 
  mutate(n = as.numeric(n))

str(sales_rep_day_of_month_sales)

sales_rep_day_of_month_sales <- sales_rep_day_of_month_sales %>% 
  pivot_wider(names_from = day, values_from = n, names_prefix = "day_" ) %>% 
  as.data.frame() %>% 
  select(year, month, day_1, day_28, day_29, day_30, day_31) %>% 
  filter(!is.na(day_1))

str(sales_rep_day_of_month_sales)
```

There are two months with multiple sales rep order days for 2011, (11/08 and 11/10), one for 2012, (1201), and two in 2014, (14/01 and 14/03).  The 14/03 is the only three day sales rep order month.

Are there months where there were no sales recorded for the sales reps?

There are two approaches.  The first is to generate a list of months between the beginning and end of history and compare that to the Sales Rep records
```{r}
monthly_sales_rep_sales <- monthly_sales_w_channel %>% 
  filter(onlineorderflag == "Sales Rep") %>% 
  mutate(orderdate = as.Date(floor_date(orderdate, "month"))) %>% 
  count(orderdate)

str(monthly_sales_rep_sales)

date_list <- tibble(month_date = seq.Date(floor_date(as.Date(min_soh_dt), "month"), 
         floor_date(as.Date(max_soh_dt), "month"), 
         by = "month"),
         date_exists = FALSE)

date_list %>% 
  anti_join(monthly_sales_rep_sales, 
            by = c("month_date" = "orderdate") )
```


  * June, September, and November are missing for 2011. 
  * June for 2014

The second approach is to use the dates found in the database for online orders.  Defining "complete" may not always be as simple as generating a complete list of months.
```{r}
sales_order_header_online <- tbl(con, in_schema("sales", "salesorderheader")) %>% 
  filter(onlineorderflag == TRUE) %>% 
  mutate(
    orderdate = date_trunc('month', orderdate)
  ) %>%
  count(orderdate, name = "online_count") 

sales_order_header_sales_rep <- tbl(con, in_schema("sales", "salesorderheader")) %>% 
  filter(onlineorderflag == FALSE) %>% 
  mutate(
    orderdate = date_trunc('month', orderdate)
  ) %>%
  count(orderdate, name = "sales_rep_count") 

missing_dates <- sales_order_header_sales_rep %>% 
  full_join(sales_order_header_online) %>% 
  show_query() %>% 
  collect()

missing_dates <- sales_order_header_online %>% 
  anti_join(sales_order_header_sales_rep) %>% 
  arrange(orderdate) %>% 
  collect()

missing_dates
str(missing_dates)
```

And in this case they agree!

discuss February issues.  and stuff. 

look at each year sepraately as a diagnostic

Use the same pivot strategy on the corrected data.

difference between detective work with a graph and just print it out.  "now I see what's driving the hint." 

We have xx months when we add the month before and the month after the **suspicious months**.  We don't know whether the problem postings have been carried forward or backward.  We check for and eliminate duplicates as well.

*  Most of the **Sales Reps**' orders are entered on a single day of the month, unique days = 1. It is possible that these are monthly recurring orders that get released on a given day of the month.  If that is the case, what are the **Sales Reps** doing the rest of the month?
*  ** ?? The lines with multiple days, unique_days > 1, have a noticeable higher number of orders, so_cnt, and associated so dollars.?? **


## Correcting the order date for **Sales Reps**


```{r}
dbGetQuery(
  con,
  "
with udays as (
SELECT to_char(orderdate,'YYMM') yymm
     ,EXTRACT(YEAR FROM soh.orderdate) yr
     , EXTRACT(MONTH FROM soh.orderdate) mo 
     , COUNT(DISTINCT soh.orderdate) *1.0 unique_days
     , COUNT(*) so_cnt
     , sum(subtotal) so_dollars
  FROM sales.salesorderheader soh
where not onlineorderflag
group by to_char(orderdate,'YYMM') 
     , EXTRACT(MONTH FROM orderdate) 
     , EXTRACT(YEAR FROM orderdate)
ORDER BY to_char(orderdate,'YYMM')
)
select soh.orderdate,count(*) from udays 
join sales.salesorderheader soh on to_char(soh.orderdate,'YYMM')  = udays.yymm
where unique_days > 1
  and not onlineorderflag
group by soh.orderdate
having count(*) > 1
order by orderdate
"
)
```

The former query misses the Sales Rep sales recorded on the first of the month in the following months:

  * 2011-12-01: 40
  * 2014-05-01: 179

### Define a date correction function in R

This code does the date-correction work on the R side:

```{r}
monthly_sales_rep_adjusted <- tbl(con, in_schema("sales", "salesorderheader")) %>%
  filter(onlineorderflag == FALSE) %>% 
  select(orderdate, subtotal, onlineorderflag) %>%
  group_by(orderdate) %>%
  summarize(
    total_soh_dollars = round(sum(subtotal, na.rm = TRUE), 2),
    soh_count = n()
  ) %>%
  mutate(
    orderdate = as.Date(orderdate),
    day = day(orderdate)
  ) %>%
  # show_query() %>%
  collect() %>%
  ungroup() %>% 
  mutate(
    adjusted_orderdate = case_when(
      day == 1L ~ orderdate -1,
      TRUE ~ orderdate
    ),
    year_month = floor_date(adjusted_orderdate, "month")
  ) %>% 
  group_by(year_month) %>% 
  summarize(
      total_soh_dollars = round(sum(total_soh_dollars, na.rm = TRUE), 2),
      soh_count = sum(soh_count)
    ) %>%
  ungroup()
```
Inspect:

```{r}
str(monthly_sales_rep_adjusted)
monthly_sales_rep_adjusted %>% filter(year(year_month) %in% c(2011,2014))
```

### Define and store a PostgreSQL function to correct the date{#define-postgres-date-function}

The following code defines a function on the server side to correct the date:

```{r}
dbExecute(
  con,
  "CREATE OR REPLACE FUNCTION so_adj_date(so_date timestamp, ONLINE_ORDER boolean) RETURNS timestamp AS $$
     BEGIN
        IF (ONLINE_ORDER) THEN
            RETURN (SELECT so_date);
        ELSE
            RETURN(SELECT CASE WHEN EXTRACT(DAY FROM so_date) = 1
                               THEN  so_date - '1 day'::interval
                               ELSE  so_date
                          END
                  );
        END IF;
 END; $$
LANGUAGE PLPGSQL;
"
)
```

### Use the PostgreSQL function

If you can do the heavy lifting on the database side, that's good.  R can do it, but it's best for finding the issues.

```{r}
monthly_sales_rep_adjusted_with_psql_function <- tbl(con, in_schema("sales", "salesorderheader")) %>%
  select(orderdate, subtotal, onlineorderflag) %>%
  mutate(
    orderdate = as.Date(orderdate)) %>%
  mutate(adjusted_orderdate = as.Date(so_adj_date(orderdate, onlineorderflag))) %>%
  filter(onlineorderflag == FALSE) %>%
  group_by(adjusted_orderdate) %>%
  summarize(
    total_soh_dollars = round(sum(subtotal, na.rm = TRUE), 2),
    soh_count = n()
  ) %>%
  # show_query() %>%
  collect() %>%
  mutate( year_month = floor_date(adjusted_orderdate, "month")) %>% 
    group_by(year_month) %>%
  ungroup() %>% 
  arrange(year_month)
```

```{r}
monthly_sales_rep_adjusted_with_psql_function %>% 
  filter(year(year_month) %in% c(2011,2014))

```

There's one minor difference between the two:
```{r}
all_equal(monthly_sales_rep_adjusted, monthly_sales_rep_adjusted_with_psql_function)
```


### Monthly Sales by Order Type with corrected dates -- relative to a trend line


```{r, eval=FALSE}

monthly_sales_rep_as_is <- monthly_sales_w_channel %>%
  filter(onlineorderflag == "Sales Rep")


ggplot(
  data = monthly_sales_rep_adjusted,
  aes(x = year_month, y = soh_count)
) +
  geom_line(alpha = .5) +
  geom_smooth(se = FALSE) +
  geom_smooth(
    data = monthly_sales_rep_as_is, aes(
      orderdate, soh_count
    ), color = "red", alpha = .5,
    se = FALSE
  ) +
  theme(plot.title = element_text(hjust = .5)) + # Center ggplot title
  labs(
    title = glue(
      "Number of Sales per month using corrected dates\n",
      "Counting Sales Order Header records"
    ),
    x = paste0("Monthly - between ", min_soh_dt, " - ", max_soh_dt),
    y = "Number of Sales Recorded"
  )
```
 
 
```{r}

monthly_sales_rep_as_is <- monthly_sales_w_channel %>%
  filter(onlineorderflag == "Sales Rep") %>% 
  mutate(orderdate = as.Date(floor_date(orderdate, unit = "month"))) %>% 
  group_by(orderdate) %>% 
  summarize(
      total_soh_dollars = round(sum(total_soh_dollars, na.rm = TRUE), 2),
      soh_count = sum(soh_count)
    ) 

monthly_sales_rep_as_is %>% 
  filter(year(orderdate) %in% c(2011,2014))


ggplot(
  data = monthly_sales_rep_adjusted,
  aes(x = year_month, y = soh_count)
) +
  geom_line(alpha = .5 , color = "green") +
  geom_point(alpha = .5 , color = "green") +
  geom_point(
    data = monthly_sales_rep_as_is, aes(
      orderdate, soh_count), color = "red", alpha = .5) +
  theme(plot.title = element_text(hjust = .5)) + # Center ggplot title
  annotate(geom = "text", y = 250, x = as.Date("2011-06-01"), 
           label = "Green dots: corrected data\nOrange dots: original data\nBrown dots: unchanged",
           hjust = 0) +
  labs(
    title = glue(
      "Number of Sales per month \nOriginal and corrected amounts"
    ),
    subtitle = glue("Subtitle"),
    caption = glue("Datasets Include: \n
                   monthly_sales_rep_adjusted, monthly_sales_rep_as_is"),
    x = paste0("Monthly - between ", min_soh_dt, " - ", max_soh_dt),
    y = "Number of Sales Recorded"
  )
```

```{r}
mon_sales <- monthly_sales_rep_adjusted %>% 
  rename(orderdate = year_month)

sales_original_and_adjusted <- bind_rows(mon_sales, monthly_sales_rep_as_is, .id = "date_kind")

```

Does this graph add anything important?

```{r}
ggplot(
  data = sales_original_and_adjusted,
  aes(x = orderdate, y = soh_count, fill = date_kind)
) +
  geom_col(position = "dodge") +
  theme(plot.title = element_text(hjust = .5)) + # Center ggplot title
  labs(
    title = glue(
      "Number of Sales per month using corrected dates\n",
      "Counting Sales Order Header records"
    ),
    subtitle = glue("Subtitle"),
    caption = glue("Datasets Include: \n
                   monthly_sales_rep_adjusted, monthly_sales_rep_as_is"),
    x = paste0("Monthly - between ", min_soh_dt, " - ", max_soh_dt),
    y = "Number of Sales Recorded",
    fill = "Date\nadjustment"
  ) 
```

additive graph showing how correction adds in some months and subtracts in others.

only show months where there is a mis-match.

```{r}
ggplot(data = monthly_sales_rep_adjusted, aes(x = year_month, y = total_soh_dollars)) +
  geom_line() +
  scale_y_continuous(labels = dollar) +
  theme(plot.title = element_text(hjust = 0.5)) + # Center the title
  labs(
    title = glue("Sales by Month - Sales Reps only\n", {format(min_soh_dt, "%B %d, %Y")} , " to  ", 
            {format(max_soh_dt, "%B %d, %Y")}),
    x = "Month - Adjusted dates",
    y = "Sales Dollars"
  )
```

Sales still seem to gyrate!

## Disconnect from the database and stop Docker

```{r}
dbDisconnect(con)
# use connection_close(con) when running interactively

#sp_docker_stop("adventureworks")
```
