# Sales Forecasting {#chapter_sales-forecasting}

> This chapter demonstrates how to:
> 
> * Use [`tidyverts`](https://tidyverts.org/) packages to explore and forecast sales data.

## Setup

The following packages are used in this chapter:
```{r package list, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(DBI)
library(RPostgres)
require(knitr)
library(bookdown)
library(sqlpetr)
library(tsibble)
library(fable)
```
Analyzing sales time series, in particular determining seasonality and forecasting future sales, is a common activty in business management. A collection of packages called `tidyverst` is designed to do this in a tidy data framework.

First, we make sure the Docker container is ready and connect to the `adventureworks` database.

```{r connect_to adventureworks}
sqlpetr::sp_docker_start("adventureworks")
con <- sqlpetr::sp_get_postgres_connection(
  user = Sys.getenv("DEFAULT_POSTGRES_USER_NAME"),
  password = Sys.getenv("DEFAULT_POSTGRES_PASSWORD"),
  dbname = "adventureworks",
  port = 5432, 
  seconds_to_test = 20, 
  connection_tab = FALSE
)
```
Next, we retrieve the "sales order header" table from the database, close the connection and stop the container.

```{r retrieve_sales_order_header}
dbExecute(con, "set search_path to sales;") 
salesorderheader_tibble <- DBI::dbReadTable(con, "salesorderheader")
DBI::dbDisconnect(con)
sqlpetr::sp_docker_stop("adventureworks")
```

## Exploring the sales data

Some assumptions:
1. The business requirement is to analyze / forecast revenue.
2. The revenue figures we care about are those in the `subtotal` column. The shipping and tax numbers are costs, not revenue.
3. The values in column `subtotal` have been converted to the common currency of the Adventureworks headquarters.

In a real-world setting, the analyst would need to validate these assumptions. Given that, our first task is to create a `tsibble` - a time series tibble - of monthly revenue figures.

### Creating a tsibble

```{r create_raw_tsibble}
monthly_tsibble <- salesorderheader_tibble %>%
  dplyr::mutate(
    origin = ifelse(onlineorderflag, "online", "sales_rep"),
    month_start = lubridate::floor_date(shipdate, unit = "months")
  ) %>%
  dplyr::group_by(origin, month_start) %>% 
    dplyr::summarize(orders = n(), total_revenue = sum(subtotal)) %>% 
  dplyr::ungroup() %>% 
  tsibble::as_tsibble(key = origin, index = month_start)

```

Note that there are two more months - June and July of 2014 - for the online data, and the revenue values are suspiciously low. We remove them for consistency.

```{r}
monthly_tsibble <- monthly_tsibble %>% dplyr::filter(month_start < '2014-06-01')

```

### Exploring the data

First, let's look at orders for online and sales representative sales:

```{r}
monthly_tsibble %>% autoplot(orders)

```

Wow! Online orders really took off in the late spring - early summer of 2013! How about revenues?

```{r}

monthly_tsibble %>% autoplot(total_revenue)

```

There's an increase, but the sales representatives always brought in more revenue than the online platform. And there's a pronounced variation in the revenue from sales representatives on a month-to-month basis.

Before moving on, let's look at revenue per order.

```{r}
monthly_tsibble %>% autoplot(total_revenue / orders)

```

For the sales representatives, there's still a month-to-month variation but the revenue per order appears to be bounded both below and above. However, the online revenue per order is decreasing.

## Forecasting
