# Introduction to DBMS queries (11)

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# These packages are called in almost every chapter of the book:
library(tidyverse)
library(DBI)
library(RPostgres)
library(glue)
library(here)
require(knitr)
library(dbplyr)
library(sqlpetr)
library(bookdown)
```
Assume that the Docker container with PostgreSQL and the dvdrental database are ready to go. 
```{r}
sp_docker_start("sql-pet")
```
Connect to the database:
```{r}
con <- sp_get_postgres_connection(user = Sys.getenv("DEFAULT_POSTGRES_USER_NAME"),
                         password = Sys.getenv("DEFAULT_POSTGRES_PASSWORD"),
                         dbname = "dvdrental",
                         seconds_to_test = 10)

```

## Downloading the data from the database

As we show later on, the database serves as a store of data and as an engine for sub-setting, joining, and doing computation.  We begin with simple extraction, or "downloading" data.

### Finding out what's there

We've already seen the simplest way of getting a list of tables in a database with `DBI` functions that  list tables and fields.  Here are the (public) tables in the database:
```{r}
DBI::dbListTables(con)
```
Here are the fields (or columns or variables) in one specific table:
```{r}
DBI::dbListFields(con, "rental")
```

Later on we'll discuss how to get more extensive data about each table and column from the database's own store of metadata.

### Downloading an entire table

There are many different methods of getting data from a DBMS, and we'll explore the different ways of controlling each one of them.

`DBI::dbReadTable` will download an entire table into an R [tibble](https://tibble.tidyverse.org/).
```{r}

rental_tibble <- DBI::dbReadTable(con, "rental") 
str(rental_tibble)
```
That's very simple, but if the table is large it may not be a good idea.

### Referencing a table for many different purposes

The `dplyr::tbl` function gives us more control over access to a table.  It embeds a connection object into the table object that dplyr uses for constructing queries and retrieving data from the DBMS.  
```{r}
rental_table <- dplyr::tbl(con, "rental")
```
Consider the structure of the connection object:
```{r}
str(rental_table)
```
Notice that the first list contains the source connection information.

It contains a list of variables in the table, among other things:
```{r}
rental_table$ops$vars
```
<<But because of lazy loading, R has not retrieved any actual data from the DBMS.  We can trigger data extraction in several ways.  Although printing `rental_table` just prints the connection object, `head` triggers a query and prints its results:

```{r}
# rental_table %>% head
```

smy
Because R is lazy and smart, R only retrieves the first 10 rows of data from the DBMS by default.  This is a key paradigm shift for those new to working R-dplyr with databases.  When working locally, R has everything in memory in an R object, (dataframe/data.table)  In the code block below, we see that nrows returns NA.  Dplyr syntax is being converted to SQL syntax and pushed down to the DBMS, but only 10 rows are fetched back your R environment.

See section **Controlling number of rows returned** to see how to tell R to quit being lazy, get to work, and return all the rows.

Here is the bug:
Printing `rental_table` in the next code block, returns all the rows, but in the R-Studio Console, it just prints the first 10 rows.

```{r}
nrow(rental_table)
rental_table
```






### Sub-setting variables


```{r}
rental_table %>% select(rental_date, return_date) %>% head
```

We won't discuss dplyr methods for sub-setting variables, deriving new ones, or sub-setting rows based on the values found in the table because they are covered well in other places, including:

  * Comprehensive reference: [https://dplyr.tidyverse.org/](https://dplyr.tidyverse.org/)
  * Good tutorial: [https://suzan.rbind.io/tags/dplyr/](https://suzan.rbind.io/tags/dplyr/) 

### Controlling number of rows returned

The `collect` function triggers the creation of a tibble and controls the number of rows that the DBMS sends to R.
```{r}
rental_table %>% collect(n = 3) %>% head
```
In this case the `collect` function triggers the execution of a query that counts the number of records in the table by `staff_id`:
```{r}
rental_table %>% 
  count(staff_id) %>% 
  collect()
```

The `collect` function affects how much is downloaded, not how many rows the DBMS needs to process the query. This query processes all of the rows in the table but only displays one row of output.
```{r}
rental_table %>% 
  count(staff_id) %>% 
  collect(n = 1)
```
### Examining dplyr's SQL query and re-using SQL code

The `show_query` function shows how dplyr is translating your query:
```{r}
rental_table %>% 
  count(staff_id) %>% 
  show_query()
```
Here is an extensive discussion of how dplyr code is translated into SQL:

* [https://dbplyr.tidyverse.org/articles/sql-translation.html](https://dbplyr.tidyverse.org/articles/sql-translation.html) 

The SQL code can be submitted directly to the DBMS with the `DBI::dbGetQuery` function:
```{r}
query_result <- DBI::dbGetQuery(con,
  'SELECT "staff_id", COUNT(*) AS "n"
   FROM "rental"
   GROUP BY "staff_id";
  ')

query_result
```

<<smy We haven't investigated this, but it looks like dplyr collect() function triggers a call simmilar to the dbGetQuery call above.  The default dplyr behavior looks like dbSendQuery() and dbFetch() model is used.>>

When you create a report to run repeated, you might want to put that query into R markdown.  That way you can also execute that SQL code in a chunk with the following header:

  {`sql, connection=con, output.var = "mydataframe"`}

```{sql, connection=con, output.var = "mydataframe"}
SELECT "staff_id", COUNT(*) AS "n"
FROM "rental"
GROUP BY "staff_id";
```
Rmarkdown will store the query result in a tibble:
```{r}
mydataframe
```

## Investigating a single table

Dealing with a large, complex database highlights the utility of specific tools in R.  We include brief examples that we find to be handy:

  + Base R structure: str
  + printing out some of the data: datatable / kable / View
  + summary statistics: summary
  + glimpse
  + skimr

### `str` - a base package workhorse

`str` is a workhorse function that lists variables, their type and a sample of the first few variable values.
```{r}
str(rental_tibble)
```

### Always just look at your data with `head`, `View` or `datadtable`

<<There is no substitute for looking at your data and R provides several ways to just browse it.  The `head` and `tail` functions help control the number of rows that are displayed.  In every-day practice you would look at more than just 6 rows, but hear we wrap `head` around the data frame: smy There is no substitute for looking at your data and R provides several ways to just browse it.  The `head` function controls the number of rows that are displayed.  Note that tail does not work against a database object.  In every-day practice you would look at more than the default 6 rows, but here we wrap `head` around the data frame: >>
```{r}
sp_print_df(head(rental_tibble, n=10))
```

### The `summary` function in base

The basic statistics that the base package `summary` provides can serve a unique diagnostic purpose in this context.  For example, the following output shows that `rental_id` is a sequential number from 1 to 16,049 with no gaps.  The same is true of `inventory_id`.  The number of NA's is a good first guess as to the number of dvd's rented out or lost on 2005-09-02 02:35:22.
```{r}
summary(rental_tibble)
```

Here are some packages that we find handy in the preliminary investigation of a database (or a problem that involves data from a database).
```{r}
```
### The `glimpse` function in the `tibble` package

The `tibble` package's `glimpse` function is a more compact version of `str`:
```{r}
glimpse(rental_tibble)
```
### The `skim` function in the `skmir` package

The `skimr` package has several functions that make it easy to examine an unknown data frame and assess what it contains. It is also extensible.
```{r}
library(skimr)

skim(rental_tibble)

wide_rental_skim <- skim_to_wide(rental_tibble)
```

## Dividing the work between R on your machine and the DBMS

They work together.

### Make the server do as much work as you can

* show_query as a first draft of SQL.  May or may not use SQL code submitted directly.

### Criteria for choosing between `dplyr` and native SQL

This probably belongs later in the book.

* performance considerations: first get the right data, then worry about performance
* Trade offs between leaving the data in PostgreSQL vs what's kept in R: 
  + browsing the data
  + larger samples and complete tables
  + using what you know to write efficient queries that do most of the work on the server

### dplyr tools

Where you place the `collect` function matters.â„
```{r}
dbDisconnect(con)
sp_docker_stop("sql-pet")
```

## Other resources

  * Benjamin S. Baumer, A Grammar for Reproducible and Painless Extract-Transform-Load Operations on Medium Data: [https://arxiv.org/pdf/1708.07073](https://arxiv.org/pdf/1708.07073) 

