# Introduction to DBMS queries (11)

> This chapter demonstrates how to:
> 
> * Get a glimpse of what tables are in the database and what fields a table contains
> * Download all or part of a table from the dbms
> * See how `dplyr` and `SQL` commands are equivalent
> * Get acquainted with some useful tools for investigating a single table
> * Begin thinking about how to divide the work between your local R session and the dbms

The following packages are used in this chapter:
```{r package list, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(DBI)
library(RPostgres)
library(dbplyr)
require(knitr)
library(bookdown)
library(sqlpetr)
```
Assume that the Docker container with PostgreSQL and the dvdrental database are ready to go. If not go back to [the previous Chapter][Build the pet-sql Docker Image]
```{r check on sql-pet}
sp_docker_start("sql-pet")
```
Connect to the database:
```{r connect to postgresql}
con <- sp_get_postgres_connection(
  user = Sys.getenv("DEFAULT_POSTGRES_USER_NAME"),
  password = Sys.getenv("DEFAULT_POSTGRES_PASSWORD"),
  dbname = "dvdrental",
  seconds_to_test = 10
)
```

## Getting data from the database

As we show later on, the database serves as a store of data and as an engine for sub-setting, joining, and computation on the data.  We begin with getting data from the dbms, or "downloading" data.

### Finding out what's there

We've already seen the simplest way of getting a list of tables in a database with `DBI` functions that list tables and fields.  Generate a vector listing the (public) tables in the database:
```{r}
tables <- DBI::dbListTables(con)
tables
```
Print a vector with all the fields (or columns or variables) in one specific table:
```{r}
DBI::dbListFields(con, "rental")
```

### Listing all the fields for all the tables

The first example, `DBI::dbListTables(con)` returned 22 tables and the second example, `DBI::dbListFields(con, "rental")` returns 7 fields.  Here we combine the two calls to return a list of tables which has a list of all the fields in the table.  The code block just shows the first two tables.

```{r combine table and column data}

table_columns <- lapply(tables, dbListFields, conn = con)

# or using purr:

table_columns <- map(tables, ~ dbListFields(.,conn = con) )

# rename each list [[1]] ... [[22]] to meaningful table name
names(table_columns) <- tables

head(table_columns)
```

Later on we'll discuss how to get more extensive data about each table and column from the database's own store of metadata using a similar technique.  As we go further the issue of scale will come up again and again: you need to be careful about how much data a call to the dbms will return, whether it's a list of tables or a table that could have millions of rows.

It's improtant to connect with people who own, generate, or are the subjects of the data.  A good chat with people who own the data, generate it, or are the subjects can generate insights and set the context for your investigation of the database. The purpose for collecting the data or circumsances where it was collected may be burried far afield in an organization, but *usually someone knows*.  The metadata discussed in a later chapter is essential but will only take you so far.

There are different ways of just **looking at the data**, which we explore below.

### Downloading an entire table

There are many different methods of getting data from a DBMS, and we'll explore the different ways of controlling each one of them.

`DBI::dbReadTable` will download an entire table into an R [tibble](https://tibble.tidyverse.org/).  
```{r}
rental_tibble <- DBI::dbReadTable(con, "rental")
str(rental_tibble)
```
That's very simple, but if the table is large it may not be a good idea, since R is designed to keep the entire table in memory.  Note that the first line of the str() output reports the total number of observations.  

### A table object that can be reused

The `dplyr::tbl` function gives us more control over access to a table by enabling  control over which columns and rows to download.  It creates  an object that might **look** like a data frame, but it's actually a list object that `dplyr` uses for constructing queries and retrieving data from the DBMS.  

```{r}
rental_table <- dplyr::tbl(con, "rental")
```

Consider the structure of the connection object:
```{r}
str(rental_table)
```

Notice that the first list contains the source connection information.  Among other things it contains a list of variables/columns in the table:

```{r}
rental_table$ops$vars
```

### Paradigm Shift: Lazy Execution

R and `dplyr` is designed to be both lazy and smart. Lazy execution affects when a donwload happens and when processing occurs on the dbms server side.  R retrieves the full data set when explicityly told to do so via the `collect` verb, otherwise it returns only 10 rows. And `dplyr` tries to get as much work done on the server side as possible before downloading anything.  All of this is a key paradigm shift for those new to working with databases using R and `dplyr`, especially if they have been working in a straight SQL environment.

**explain why this matters here**

In the code blocks below, we demonstrate three ways to check if `dplyr` has performed lazy/delayed execution.

1.  Display the object.
2.  nrow(object)
3.  Check the pipe.

```{r Lazy Execution 1 View Output}
rental_table <- dplyr::tbl(con, "rental")
rental_table
head(rental_table, n = 25)
```
At the top of the output look for the dimensions, [?? x columns] or at the bottom for '... with more rows'

Notice that `head` should return 25 rows, but only shows the 10 rows it has can retrieve. 

```{r Lazy Execution 2 nrow of the object}
nrow(rental_table)
```

The the `rental_table` is smart enough to returns `NA` when passed to `nrow` so no execution is performed.

```{r Lazy Execution 3 summarise implies "collect"}
rental_table %>% dplyr::summarise(n = n())
rental_table %>% dplyr::summarise(n = n()) %>% show_query()
```
The `dplyr::summarise` verb does not force R to download the whole table.  It **processes** the whole table by counting all the rows on the dbms side and then downloads one number.

Next we give an example where R gets busy and returns the full data set.

```{r}
rental_table <- dplyr::tbl(con, "rental") # Lazy

collect_rental_table <- rental_table %>% collect() # Busy
collect_rental_table
```

At the top of the output, `# A tibble: 16,044 x 7` and at the bottom of the output, `# ... with 16,034 more rows` we see that `dplyr::tbl` returned all the rows associated with `collect_rental_table`.  See [Controlling number of rows returned] for additional R examples showing R getting to work, and return some or all the rows.

See more example of lazy execution can be found [Here](https://datacarpentry.org/R-ecology-lesson/05-r-and-databases.html).


### Controlling the number of rows returned

The `collect` function triggers the creation of a tibble and controls the number of rows that the DBMS sends to R.  Note that in the following examples, the object dimensions are known.
```{r}
rental_table %>% collect(n = 3) %>% head()
```
In this case the `collect` function triggers the execution of a query that counts the number of records in the table by `staff_id`:
```{r}
rental_table %>%
  count(staff_id) %>%
  collect()
```

The `collect` function affects how much is downloaded, not how many rows the DBMS needs to process the query. This query processes all of the rows in the table but only displays one row of output.
```{r}
rental_table %>%
  count(staff_id) %>%
  collect(n = 1)
```

### Random rows from the dbms

When the dbms contains many rows, a sample of the data may be plenty for your purposes.  Although `dplyr` has nice functions to sample a data frame that's already in R (e.g., the `sample_n` and `sample_frac` functions), to get a sample from the dbms we have to use `dbGetQuery` to send native SQL to the database. To peak ahead, here is one example of a query that retrieves 20 rows from a 1% sample:

```{r}
one_percent_sample <- DBI::dbGetQuery(
  con,
  "SELECT rental_id, rental_date, inventory_id, customer_id FROM rental TABLESAMPLE SYSTEM(1) LIMIT 20;
  "
)

one_percent_sample
```

### Sub-setting variables

A table in the dbms may not only have many more rows than you want and also many more columns.  The `select` command controls which columns are retrieved.
```{r}
rental_table %>% select(rental_date, return_date) %>% head()
```

We won't discuss `dplyr` methods for sub-setting variables, deriving new ones, or sub-setting rows based on the values found in the table because they are covered well in other places, including:

  * Comprehensive reference: [https://dplyr.tidyverse.org/](https://dplyr.tidyverse.org/)
  * Good tutorial: [https://suzan.rbind.io/tags/dplyr/](https://suzan.rbind.io/tags/dplyr/) 

In practice we find that, **renaming variables** is often quite important because the names in an SQL database might not meet your needs as an analyst.  In "the wild" you will find names that are ambiguous or overly specified, with spaces in them, and other problems that will make them difficult to use in R.  It is good practice to do whatever renaming you are going to do in a predictable place like at the top of your code.  The names in the `dvdrental` database are simple and clear, but if they were not, you might rename them for subsequent use in this way:

```{r}
renamed_rental_table <- dplyr::tbl(con, "rental") %>%
  rename(rental_id_number = rental_id, inventory_id_number = inventory_id)

renamed_rental_table %>%
  select(rental_id_number, rental_date, inventory_id_number) %>%
  head()
```

### Examining `dplyr`'s SQL query and re-using SQL code

The `show_query` function shows how `dplyr` is translating your query to the dialect of the target dbms:
```{r}
rental_table %>%
  count(staff_id) %>%
  show_query()
```
Here is an extensive discussion of how `dplyr` code is translated into SQL:

* [https://dbplyr.tidyverse.org/articles/sql-translation.html](https://dbplyr.tidyverse.org/articles/sql-translation.html) 

The SQL code can submit the same query directly to the DBMS with the `DBI::dbGetQuery` function:
```{r}
DBI::dbGetQuery(
  con,
  'SELECT "staff_id", COUNT(*) AS "n"
   FROM "rental"
   GROUP BY "staff_id";
  '
)
```
<<smy We haven't investigated this, but it looks like `dplyr` collect() function triggers a call simmilar to the dbGetQuery call above.  The default `dplyr` behavior looks like dbSendQuery() and dbFetch() model is used.>>

When you create a report to run repeatedly, you might want to put that query into R markdown.  That way you can also execute that SQL code in a chunk with the following header:

  {`sql, connection=con, output.var = "query_results"`}

```{sql, connection=con, output.var = "query_results"}
SELECT "staff_id", COUNT(*) AS "n"
FROM "rental"
GROUP BY "staff_id";
```
Rmarkdown stored that query result in a tibble:
```{r}
query_results
```

## Examining a single table with R

Dealing with a large, complex database highlights the utility of specific tools in R.  We include brief examples that we find to be handy:

  + Base R structure: `str`
  + printing out some of the data: `datatable`, `kable`, and `View`
  + summary statistics: `summary`
  + `glimpse` oin the   `tibble` package, which is included in the `tidyverse`
  + `skim` in the `skimr` package

### `str` - a base package workhorse

`str` is a workhorse function that lists variables, their type and a sample of the first few variable values.
```{r}
str(rental_tibble)
```

### Always just **look** at your data with `head`, `View`, or `kable`

There is no substitute for looking at your data and R provides several ways to just browse it.  The `head` function controls the number of rows that are displayed.  Note that tail does not work against a database object.  In every-day practice you would look at more than the default 6 rows, but here we wrap `head` around the data frame: 
```{r}
sp_print_df(head(rental_tibble))
```

### The `summary` function in base

The basic statistics that the base package `summary` provides can serve a unique diagnostic purpose in this context.  For example, the following output shows that `rental_id` is a sequential number from 1 to 16,049 with no gaps.  The same is true of `inventory_id`.  The number of NA's is a good first guess as to the number of dvd's rented out or lost on 2005-09-02 02:35:22.
```{r}
summary(rental_tibble)
```

### The `glimpse` function in the `tibble` package

The `tibble` package's `glimpse` function is a more compact version of `str`:
```{r}
tibble::glimpse(rental_tibble)
```
### The `skim` function in the `skimr` package

The `skimr` package has several functions that make it easy to examine an unknown data frame and assess what it contains. It is also extensible.
```{r}
library(skimr)

skim(rental_tibble)

wide_rental_skim <- skim_to_wide(rental_tibble)
```

## Dividing the work between R on your machine and the DBMS

They work together.

### Make the server do as much work as you can

* show_query as a first draft of SQL.  May or may not use SQL code submitted directly.

### Criteria for choosing between `dplyr` and native SQL

This probably belongs later in the book.

* performance considerations: first get the right data, then worry about performance
* Trade offs between leaving the data in PostgreSQL vs what's kept in R: 
  + browsing the data
  + larger samples and complete tables
  + using what you know to write efficient queries that do most of the work on the server

### `dplyr` tools

Where you place the `collect` function matters.
```{r}
dbDisconnect(con)
sp_docker_stop("sql-pet")
```

## DBI Package

In this chapter we touched on a number of functions from the DBI Package.  The table below shows other functions in the package.  The Chapter column references a section in the book if we have used it.

|DBI| Chapter|Call Example/Notes
|---|--------|----------------------------------------
|DBIConnection-class||
|dbAppendTable||
|dbCreateTable||
|dbDisconnect|Every Chapter|
|dbExecute|10.4.2|Executes a statement and returns the number of rows affected. dbExecute() comes with a default implementation (which should work with most backends) that calls dbSendStatement(), then dbGetRowsAffected(), ensuring that the result is always free-d by dbClearResult().
|dbExistsTable||dbExistsTable(con,'actor')
|dbFetch|17.1|dbFecth(rs)
|dbGetException||
|dbGetInfo||dbGetInfo(con)
|dbGetQuery|10.4.1|dbGetQuery(con,'select * from store;')
|dbIsReadOnly||dbIsReadOnly(con)
|dbIsValid||dbIsValid(con)
|dbListFields|8.1.1|DBI::dbListFields(con, "rental")
|dbListObjects||dbListObjects(con)
|dbListResults||deprecated
|dbListTables|8.1.1|DBI::dbListTables(con, "rental")
|dbReadTable|8.1.2|DBI::dbReadTable(con, "rental")
|dbRemoveTable||
|dbSendQuery|17.1|rs <- dbSendQuery(con, "SELECT * FROM mtcars WHERE cyl = 4")
|dbSendStatement||The dbSendStatement() method only submits and synchronously executes the SQL data manipulation statement (e.g., UPDATE, DELETE, INSERT INTO, DROP TABLE, ...) to the database engine. 
|dbWriteTable|17.1|dbWriteTable(con, "mtcars", mtcars, overwrite = TRUE)


## Other resources

  * Benjamin S. Baumer, A Grammar for Reproducible and Painless Extract-Transform-Load Operations on Medium Data: [https://arxiv.org/pdf/1708.07073](https://arxiv.org/pdf/1708.07073) 

