# Joins and complex queries (13)

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# These packages are called in almost every chapter of the book:
library(tidyverse)
library(DBI)
library(RPostgres)
library(glue)
library(here)
require(knitr)
library(dbplyr)
library(sqlpetr)
```

Verify Docker is up and running:
```{r}
sp_check_that_docker_is_up()
```
verify pet DB is available, it may be stopped.
```{r}
sp_show_all_docker_containers()
```
Start up the `docker-pet` container
```{r}
sp_docker_start("sql-pet")

```


now connect to the database with R
```{r}
# need to wait for Docker & Postgres to come up before connecting.

con <- sp_get_postgres_connection(user = Sys.getenv("DEFAULT_POSTGRES_USER_NAME"),
                         password = Sys.getenv("DEFAULT_POSTGRES_PASSWORD"),
                         dbname = "dvdrental",
                         seconds_to_test = 10)


```

## Different strategies for interacting with the database


select examples
    dbGetQuery returns the entire result set as a data frame.  
        For large returned datasets, complex or inefficient SQL statements, this may take a 
        long time.

      dbSendQuery: parses, compiles, creates the optimized execution plan.  
          dbFetch: Execute optimzed execution plan and return the dataset.
    dbClearResult: remove pending query results from the database to your R environment

## Use dbGetQuery

How many customers are there in the DVD Rental System
```{r}
rs1 <- dbGetQuery(con, 'select * from customer;')
sp_print_df(head(rs1))

pco <- dbSendQuery(con, 'select * from customer;')
rs2  <- dbFetch(pco)
dbClearResult(pco)
sp_print_df(head(rs2))
```

## Use dbExecute
```{r}
# insert yourself as a new customer
dbExecute(con,
  "insert into customer 
  (store_id,first_name,last_name,email,address_id
  ,activebool,create_date,last_update,active)
  values(2,'Sophie','Yang','dodreamdo@yahoo.com',1,TRUE,'2018-09-13','2018-09-13',1)
  returning customer_id;
  "
  )

```

```{r}
## anti join -- Find customers who have never rented a movie.

rs <- dbGetQuery(con,
                 "select c.first_name
                        ,c.last_name
                        ,c.email
                    from customer c 
                         left outer join rental r
                              on c.customer_id = r.customer_id 
                   where r.rental_id is null;
                 "
                 )
sp_print_df(head(rs))
```

```{r }
## how many films and languages exist in the DVD rental application
rs <- dbGetQuery(con,
                "      select 'film' table_name,count(*) count from film 
                 union select 'language' table_name,count(*) count from language 
               ;
                "
                )
sp_print_df(head(rs))
## what is the film distribution based on language

rs <- dbGetQuery(con,
                "select l.language_id id
                       ,l.name
                       ,sum(case when f.language_id is not null then 1 else 0 end) total
                   from language l
                        full outer join film f
                             on l.language_id = f.language_id
                  group by l.language_id,l.name 
                  order by l.name;
                 ;
                "
                )
sp_print_df(head(rs))
```

```{r}
## Store analysis
### which store has had more rentals and income
rs <- dbGetQuery(con,
                "select *
                 from (      select 'actor' tbl_name,count(*) from actor 
                       union select 'category' tbl_name,count(*) from category
                       union select 'film' tbl_name,count(*) from film
                       union select 'film_actor' tbl_name,count(*) from film_actor
                       union select 'film_category' tbl_name,count(*) from film_category
                       union select 'language' tbl_name,count(*) from language
                       union select 'inventory' tbl_name,count(*) from inventory
                       union select 'rental' tbl_name,count(*) from rental
                       union select 'payment' tbl_name,count(*) from payment
                       union select 'staff' tbl_name,count(*) from staff
                       union select 'customer' tbl_name,count(*) from customer
                       union select 'address' tbl_name,count(*) from address
                       union select 'city' tbl_name,count(*) from city
                       union select 'country' tbl_name,count(*) from country
                       union select 'store' tbl_name,count(*) from store
                       ) counts
                  order by tbl_name
                 ;
                "
                )
sp_print_df(head(rs))
```

```{r}
## Store analysis
### which store has the largest income stream
rs <- dbGetQuery(con,
                "select store_id,sum(amount) amt,count(*) cnt 
                   from payment p 
                        join staff s 
                          on p.staff_id = s.staff_id  
                 group by store_id order by 2 desc
                 ;
                "
                )
sp_print_df(head(rs))
```

```{r}
## Store analysis
### How many rentals have not been paid
### How many rentals have been paid
### How much has been paid
### What is the average price/movie
### Estimate the outstanding balance
rs <- dbGetQuery(con,
                "select sum(case when payment_id is null then 1 else 0 end) missing
                       ,sum(case when payment_id is not null then 1 else 0 end) found
                       ,sum(p.amount) amt
                       ,count(*) cnt 
                       ,round(sum(p.amount)/sum(case when payment_id is not null then 1 else 0 end),2) avg_price
                       ,round(round(sum(p.amount)/sum(case when payment_id is not null then 1 else 0 end),2)
                                  * sum(case when payment_id is null then 1 else 0 end),2) est_balance
                   from rental r 
                        left outer join payment p 
                          on r.rental_id = p.rental_id  
                 ;
                "
                )
sp_print_df(head(rs))
```

```{r}
### what is the actual outstanding balance

rs <- dbGetQuery(con,
                "select sum(f.rental_rate) open_amt,count(*) count
                   from rental r 
                        left outer join payment p 
                          on r.rental_id = p.rental_id  
                        join inventory i
                          on r.inventory_id = i.inventory_id
                        join film f
                          on i.film_id = f.film_id
                  where p.rental_id is null
                 ;"
                )  
sp_print_df(head(rs))
```

```{r}
### Rank customers with highest open amounts

rs <- dbGetQuery(con,
                "select c.customer_id,c.first_name,c.last_name,sum(f.rental_rate) open_amt,count(*) count
                   from rental r 
                        left outer join payment p 
                          on r.rental_id = p.rental_id  
                        join inventory i
                          on r.inventory_id = i.inventory_id
                        join film f
                          on i.film_id = f.film_id
                        join customer c
                          on r.customer_id = c.customer_id
                  where p.rental_id is null
                  group by c.customer_id,c.first_name,c.last_name
                  order by open_amt desc
                  limit 25
                 ;"
                )  
sp_print_df(head(rs))
```

```{r}
### what film has been rented the most
rs <- dbGetQuery(con,
                "select i.film_id,f.title,rental_rate,sum(rental_rate) revenue,count(*) count  --16044
                   from rental r 
                        join inventory i
                          on r.inventory_id = i.inventory_id
                        join film f
                          on i.film_id = f.film_id
                 group by i.film_id,f.title,rental_rate
                 order by count desc
                 ;"
                )  
sp_print_df(head(rs))
```

```{r}
### what film has been generated the most revenue assuming all amounts are collected
rs <- dbGetQuery(con,
                "select i.film_id,f.title,rental_rate
                       ,sum(rental_rate) revenue,count(*) count  --16044
                   from rental r 
                        join inventory i
                          on r.inventory_id = i.inventory_id
                        join film f
                          on i.film_id = f.film_id
                 group by i.film_id,f.title,rental_rate
                 order by revenue desc
                 ;"
                )  
sp_print_df(head(rs))
```

```{r}
### which films are in one store but not the other.
rs <- dbGetQuery(con,
                "select coalesce(i1.film_id,i2.film_id) film_id
                       ,f.title,f.rental_rate,i1.store_id,i1.count,i2.store_id,i2.count
                   from     (select film_id,store_id,count(*) count 
                               from inventory where store_id = 1 
                             group by film_id,store_id) as i1
                         full outer join 
                            (select film_id,store_id,count(*) count
                               from inventory where store_id = 2 
                             group by film_id,store_id
                            ) as i2
                           on i1.film_id = i2.film_id 
                         join film f 
                           on coalesce(i1.film_id,i2.film_id) = f.film_id
                  where i1.film_id is null or i2.film_id is null 
                 order by f.title  ;
               "
                )  
sp_print_df(head(rs))
```

```{r}
# Compute the outstanding balance.
rs <- dbGetQuery(con,
                "select sum(f.rental_rate) open_amt,count(*) count
                   from rental r 
                        left outer join payment p 
                          on r.rental_id = p.rental_id  
                        join inventory i
                          on r.inventory_id = i.inventory_id
                        join film f
                          on i.film_id = f.film_id
                  where p.rental_id is null
                 ;"
                )  
sp_print_df(head(rs))
```
list what's there
```{r}
dbListTables(con)
```

Clean up
```{r}
# dbRemoveTable(con, "cars")
# dbRemoveTable(con, "mtcars")
# dbRemoveTable(con, "cust_movies")

# diconnect from the db
dbDisconnect(con)

sp_docker_stop("sql-pet")

```

