# Writing to the DBMS (73)

At the end of this chapter, you will be able to 

  * Write queries in R using docker container.
  * Start and connect to the database with R.
  * Create, Modify, and remove the table.

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# These packages are called in almost every chapter of the book:
library(tidyverse)
library(DBI)
library(RPostgres)
library(glue)
library(here)
require(knitr)
library(dbplyr)
library(sqlpetr)
```
Start up the `docker-pet` container:

```{r}
sp_docker_start("sql-pet")


```


Now connect to the database with R using your login info:

```{r}

con <- sp_get_postgres_connection(user = Sys.getenv("DEFAULT_POSTGRES_USER_NAME"),
                         password = Sys.getenv("DEFAULT_POSTGRES_PASSWORD"),
                         dbname = "dvdrental",
                         seconds_to_test = 10)
```
## Create a new table

This is an example from the DBI help file.

```{r}
dbWriteTable(con, "cars", head(cars, 3)) # "cars" is a built-in dataset, not to be confused with mtcars

dbReadTable(con, "cars")   # there are 3 rows

```
## Modify an existing table

To add additional rows or instances to the "cars" table, we will use INSERT command with their values.

There are two different ways of adding values: list them or pass values using the param argument. 

```{r}
dbExecute(
  con,
  "INSERT INTO cars (speed, dist) VALUES (1, 1), (2, 2), (3, 3)"
)
dbReadTable(con, "cars")   # there are now 6 rows

# Pass values using the param argument:
dbExecute(
  con,
  "INSERT INTO cars (speed, dist) VALUES ($1, $2)",
  param = list(4:7, 5:8)
)
dbReadTable(con, "cars")   # there are now 10 rows

```

## Remove table and Clean up

Here you will remove the table "cars", disconnect from the database and exit docker.

```{r}
dbRemoveTable(con, "cars")

# diconnect from the db
dbDisconnect(con)

sp_docker_stop("sql-pet")
```

