# Creating the sql-pet Docker container one step at a time {#chapter_step-at-a-time-docker}

> This will be the basis for a vignette in the sqlpetr package.

Step-by-step Docker container setup with dvdrental database installed
This needs to run *outside a project* to compile correctly because of
the complexities of how knitr sets working directories (or because we
don’t really understand how it works!) The purpose of this code is to

  - Replicate the docker container generated in Chapter 5 of the book,
    but in a step-by-step fashion
  - Show that the `dvdrental` database persists when stopped and started
    up again.

## Overview

Doing all of this in a step-by-step way that might be useful to
understand how each of the steps involved in setting up a persistent
PostgreSQL database works. If you are satisfied with the method shown in
Chapter 5, skip this and only come back if you’re interested in picking
apart the steps.

```{r}
library(tidyverse)
library(DBI)
library(RPostgres)
library(glue)
require(knitr)
library(dbplyr)
library(sqlpetr)
library(here)
```

## Download the `dvdrental` backup file

The first step is to get a local copy of the `dvdrental` PostgreSQL
**restore file**. It comes in a zip format and needs to be un-zipped.

```{r}
opts_knit$set(root.dir = normalizePath('../'))
if (!require(downloader)) install.packages("downloader")
library(downloader)

download("http://www.postgresqltutorial.com/wp-content/uploads/2017/10/dvdrental.zip", destfile = glue(here("dvdrental.zip")))

unzip("dvdrental.zip", exdir = here()) # creates a tar archhive named "dvdrental.tar"
```

Check on where we are and what we have in this directory:

```{r}

dir(path = here(), pattern = "^dvdrental(.tar|.zip)")

sp_show_all_docker_containers()

```

Remove the `sql-pet` container if it exists (e.g., from a prior run)

```{r}

if (system2("docker", "ps -a", stdout = TRUE) %>%
    grepl(x = ., pattern = 'sql-pet') %>%
    any()) {
  sp_docker_remove_container("sql-pet")
}
```

## Build the Docker Container

Build an image that derives from postgres:10. Connect the local and
Docker directories that need to be shared. Expose the standard
PostgreSQL port 5432.

```{r}
wd <- here()
wd

docker_cmd <- glue(
  "run ",      # Run is the Docker command.  Everything that follows are `run` parameters.
  "--detach ", # (or `-d`) tells Docker to disconnect from the terminal / program issuing the command
  " --name sql-pet ",     # tells Docker to give the container a name: `sql-pet`
  "--publish 5432:5432 ", # tells Docker to expose the PostgreSQL port 5432 to the local network with 5432
  "--mount ", # tells Docker to mount a volume -- mapping Docker's internal file structure to the host file structure
  'type=bind,source="', wd, '",target=/petdir',
  " postgres:10 " # tells Docker the image that is to be run (after downloading if necessary)
)

docker_cmd
```



```{r}
system2("docker", docker_cmd, stdout = TRUE, stderr = TRUE)
```

Peek inside the docker container and list the files in the `petdir`
directory. Notice that `dvdrental.tar` is in both.

```{r}
# local file system:
dir(path = here(), pattern = "^dvdrental.tar")

# inside docker
system2('docker', 'exec sql-pet ls petdir | grep "dvdrental.tar" ',
        stdout = TRUE, stderr = TRUE)
Sys.sleep(3)
```


## Create the database and restore from the backup

We can execute programs inside the Docker container with the `exec`
command. In this case we tell Docker to execute the `psql` program
inside the `sql-pet` container and pass it some commands as follows.

```{r}
sp_show_all_docker_containers()
```
inside Docker, execute the postgress SQL command-line program to create the dvdrental database:

```{r}
system2('docker', 'exec sql-pet psql -U postgres -c "CREATE DATABASE dvdrental;"',
        stdout = TRUE, stderr = TRUE)
Sys.sleep(3)
```

The `psql` program repeats back to us what it has done, e.g., to create
a database named `dvdrental`. Next we execute a different program in the
Docker container, `pg_restore`, and tell it where the restore file is
located. If successful, the `pg_restore` just responds with a very
laconic `character(0)`. 

```{r create the database inside Docker}

system2("docker", "exec sql-pet pg_restore -U postgres -d dvdrental petdir/dvdrental.tar", stdout = TRUE, stderr = TRUE)
Sys.sleep(3)

```

## Connect to the database with R

If you are interested take a look inside the `sp_get_postgres_connection` function to see how the DBI package is beingcused.

```{r}
con <- sp_get_postgres_connection(user = Sys.getenv("DEFAULT_POSTGRES_USER_NAME"),
                                  password = Sys.getenv("DEFAULT_POSTGRES_PASSWORD"),
                                  dbname = "dvdrental",
                                  seconds_to_test = 30)

dbListTables(con)
```

```{r}
dbDisconnect(con)

# Stop and start to demonstrate persistence
```

Stop the container

```{r}
sp_docker_stop("sql-pet")
```

Restart the container and verify that the dvdrental tables are still
there

```{r}
sp_docker_start("sql-pet")

con <- sp_get_postgres_connection(user = Sys.getenv("DEFAULT_POSTGRES_USER_NAME"),
                                  password = Sys.getenv("DEFAULT_POSTGRES_PASSWORD"),
                                  dbname = "dvdrental",
                                  seconds_to_test = 30)
dbListTables(con)
```

## Cleaning up

It’s always good to have R disconnect from the database

```{r}
dbDisconnect(con)
```

Stop the container and show that the container is still there, so can be
started again.

```{r}
sp_docker_stop("sql-pet")
```

show that the container still exists even though it’s not running

```{r}
sp_show_all_docker_containers()

```

We are leaving the `sql-pet` container intact so it can be used in running the
rest of the examples and book. 

Clean up by removing the local files used in creating the database:
```{r}
file.remove(here("dvdrental.zip"))
file.remove(here("dvdrental.tar"))

```

