---
title: "Potential Architectures"
author: "M, Edward (Ed) Borasky"
date: "9/14/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require(DiagrammeR)) install.packages("DiagrammeR")
library(DiagrammeR)
```

## Small architecture
The simplest architecture we can possibly use has just one container, running PostgreSQL. 

* We talk to the PostgreSQL container for data analysis from RStudio on the host, using the `DBI` and `RPostgres` packages.
* We talk to the PostgreSQL container for administration by building `docker exec` commands and executing them with `system2`.
* We either mount the `Backups` volume on the host filesystem or we copy files to and from `Backups` with `docker cp` commands wrapped with `system2`.

```{r echo=FALSE}
DiagrammeR::mermaid("
graph LR
    Host_Filesystem---Backups
    RStudio---PostgreSQL
    subgraph Containers
    PostgreSQL
    end
    subgraph Volumes
    PostgreSQL---pgdata
    PostgreSQL---Backups
    end
end
")
```

<br><br><br>

## Medium architecture
The medium architecture adds a `pgAdmin4` container for administering the PostgreSQL server. We have the same workflow for backups, and we still do the data analysis with host RStudio, but we manage the server with a browser pointed at the `pgAdmin4` web service.

```{r echo=FALSE}
DiagrammeR::mermaid("
graph LR
    Host_Filesystem---Backups
    RStudio---PostgreSQL
    Browser---pgAdmin4
    subgraph Containers
    PostgreSQL---pgAdmin4
    end
    subgraph Volumes
    pgAdmin4---Backups
    PostgreSQL---pgdata
    end
end
")
```

<br><br><br>

## Large architecture
In the large architecture, we add a `rocker/rstudio` container, thus creating a fully-containerized workflow. We talk to the containers via a browser only.

```{r echo=FALSE}
DiagrammeR::mermaid("
graph LR
    Host_Filesystem---Backups
    Browser---Rocker_RStudio
    Browser---pgAdmin4
    subgraph Containers
    PostgreSQL---pgAdmin4
    Rocker_RStudio---PostgreSQL
    end
    subgraph Volumes
    pgAdmin4---Backups
    PostgreSQL---pgdata
    end
end")
```
