---
title: "Set-up and connect to adventureworks database"
author: "John D. Smith"
date: "6/19/2019"
output: html_document
---
Following `https://github.com/lorint/AdventureWorks-for-Postgres`

https://docs.microsoft.com/en-us/previous-versions/sql/sql-server-2008/ms124438(v=sql.100) 

https://docs.microsoft.com/en-us/previous-versions/sql/sql-server-2008/ms124917%28v%3dsql.100%29
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Libraries
```{r load R packages}
library(tidyverse)
library(DBI)
library(RPostgres)
library(glue)
library(here)
require(knitr)
library(dbplyr)
library(sqlpetr)
library(inspectdf)
wd <- here()

sp_docker_remove_container("adventureworks")
```

download:

* `https://github.com/Microsoft/sql-server-samples/releases/download/adventureworks/AdventureWorks-oltp-install-script.zip`
* `https://github.com/lorint/AdventureWorks-for-Postgres` 

Move to `./adventureworks` directory.  Unzip `AdventureWorks-for-Postgres-master.zip`

run:
  `$ ruby update_csvs.rb`  # cleans up some of the csv files that have Microsoft junk.


```{r}

docker_cmd <- glue(
  "run ",      # Run is the Docker command.  Everything that follows are `run` parameters.
  "--detach ", # (or `-d`) tells Docker to disconnect from the terminal / program issuing the command
  " --name adventureworks ",     # tells Docker to give the container a name: `sql-pet`
  "--publish 5432:5432 ", # tells Docker to expose the PostgreSQL port 5432 to the local network with 5432
  "--mount ", # tells Docker to mount a volume -- mapping Docker's internal file structure to the host file structure
  'type=bind,source="', wd, '",target=/petdir',
  " postgres:10 " # tells Docker the image that is to be run (after downloading if necessary)
)
cat('docker ',docker_cmd)

system2("docker", docker_cmd, stdout = TRUE, stderr = TRUE)

```
Build the adventureworks database
```{r}
system2('docker', 
        'exec adventureworks psql -U postgres -c "CREATE DATABASE adventureworks;"',
        stdout = TRUE, stderr = TRUE)

```
Run:

  docker exec -ti adventureworks bash
  
Then:
  \cd petdir/adventureworks/
  
  psql -U postgres -d adventureworks 
  
  \i install.sql
  \dn 


```{r}

sp_docker_start("adventureworks")

con <- sp_get_postgres_connection(
  host = "localhost",
  port = 5432,
  user = "postgres",
  password = "postgres",
  dbname = "adventureworks",
  seconds_to_test = 10
)
```

```{r}

adventureworks_meta_data <- tbl(con, in_schema("information_schema", "columns")) %>% 
  collect() %>% 
  filter(!str_detect(table_schema,"pg_|information_schema" ))

cat_info <- inspect_cat(adventureworks_meta_data)
cat_info %>% show_plot()

# select the most useful columns
adventureworks_meta_data <- adventureworks_meta_data %>% 
  select(table_schema, table_name, column_name, dtd_identifier, data_type, column_default)

adventureworks_meta_data %>% 
  filter(table_name == "stateprovince") %>% 
  select(column_name, dtd_identifier, data_type, column_default)

stateprovince <- tbl(con, in_schema("person", "stateprovince")) %>% 
  collect()

stateprovince

index_list <- tbl(con,  "pg_indexes") %>%
  select(tablename, indexname, indexdef) %>%
  arrange(tablename) %>%
  collect() %>%
  filter(!str_starts(tablename, "pg_")) 
index_list

```
Merge two tables
```{r}

employee <- tbl(con, dbplyr::in_schema("humanresources", "employee"))

department_history  <- tbl(con, dbplyr::in_schema("humanresources", "employeedepartmenthistory"))

employee %>% 
  left_join(department_history, by = c("businessentityid") ) %>% 
  count(businessentityid, sort = TRUE)

```

```{r}
DBI::dbDisconnect(con)
```

Create the compressed PostgreSQL backup file
```{r}
docker_cmd <- glue(
  "exec  adventureworks  pg_dump adventureworks -U postgres -Fc > book-src/adventureworks.sql"
)

system2("docker", docker_cmd, stdout = TRUE, stderr = TRUE)
```

Clean up:
```{r}
system2("docker", "stop adventureworks", stdout = TRUE, stderr = TRUE)

sp_docker_remove_container("adventureworks")
```

