---
title: "AW SO EDA - # 2"
author: "Sophie Yang"
date: "July 19, 2019"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=10)
```

```{r libraries, include=FALSE}
library(tidyverse)
library(DBI)
library(RPostgres)
library(glue)
require(knitr)
library(dbplyr)
library(sqlpetr)
library(bookdown)
library(here)

library(scales)  # ggplot xy scales
```

```{r, start adventureworks}
sp_docker_start('adventureworks')
# Sys.sleep(10)
con <- sp_get_postgres_connection(
  host = "localhost",
  port = 5432,
  user = "postgres",
  password = "postgres",
  dbname = "adventureworks",
  seconds_to_test = 60, connection_tab = TRUE
)
```

# Online Orders Versus Sales Rep Orders

In a previous section, we examined how many orders are placed during the month, the associated dollars associated with online orders versus sales rep orders.  In the next code block we look at the number of distinct days associated with the online versus sales rep orders.

```{r fig.width=15}
mo_so_mo_dt_dist_sum <- dbGetQuery(con,
"
SELECT to_char(orderdate,'YYMM') yymm
     ,EXTRACT(YEAR FROM soh.orderdate) yr
     , EXTRACT(MONTH FROM soh.orderdate) mo 
--     , EXTRACT(DAY FROM soh.orderdate) dd 
     , case when onlineorderflag then 'online' else 'sales rep' end so_type
     , COUNT(DISTINCT soh.orderdate) *1.0 unique_days
     , COUNT(*) so_cnt
     , sum(subtotal) so_dollars
  FROM sales.salesorderheader soh
where 1 = 1 --not onlineorderflag
group by to_char(orderdate,'YYMM') 
     , EXTRACT(MONTH FROM orderdate) 
     , EXTRACT(YEAR FROM orderdate)
     , case when onlineorderflag then 'online' else 'sales rep' end 
--     , soh.orderdate
ORDER BY to_char(orderdate,'YYMM')
")

mo_so_mo_dt_dist_sum

ggplot(data=mo_so_mo_dt_dist_sum,aes(x=yymm,y=unique_days,fill=so_type)) +
  geom_col(position='dodge') +
  xlab("YYMM") +
  ylab("Unique Order Days in Month") +
  ggtitle("Sale Rep Order Days By Month") +
  theme(plot.title = element_text(hjust = 0.5)) +           # Center ggplot title
  theme(axis.text.x=element_text(angle=60, hjust=1))
```

The online orders are placed through out the month as one would expect.  However, there are a number of months where the company had a 0 on-line order day.  Are these associated with possible maintenance or scheduled/unscheduled downtimes.  What is going on with the salesreps?  They are placing orders on less than five days out of each month.  

# Sales Rep Orderdate Distribution

```{r fig.width=15}
mo_so_mo_dt_dist_sum <- dbGetQuery(con,
"
SELECT to_char(orderdate,'YYMM') yymm
     ,EXTRACT(YEAR FROM soh.orderdate) yr
     , EXTRACT(MONTH FROM soh.orderdate) mo 
--     , EXTRACT(DAY FROM soh.orderdate) dd 
     , COUNT(DISTINCT soh.orderdate) *1.0 unique_days
     , COUNT(*) so_cnt
     , sum(subtotal) so_dollars
  FROM sales.salesorderheader soh
where not onlineorderflag
group by to_char(orderdate,'YYMM') 
     , EXTRACT(MONTH FROM orderdate) 
     , EXTRACT(YEAR FROM orderdate)
--     , soh.orderdate
ORDER BY to_char(orderdate,'YYMM')
")

mo_so_mo_dt_dist_sum

ggplot(data=mo_so_mo_dt_dist_sum,aes(x=yymm,y=unique_days)) +
  geom_col(fill = 'blue') +
  xlab("YYMM") +
  ylab("Unique Order Days in Month") +
  ggtitle("Sale Rep Order Days By Month") +
  theme(plot.title = element_text(hjust = 0.5)) +           # Center ggplot title
  theme(axis.text.x=element_text(angle=60, hjust=1))
```

That is unexpected.  A couple of  things immediately jump out from the first page of data:

*  July, September, and November are missing for 2011. 
*  Most of the sales reps' orders are entered on a single day of the month, unique days = 1. It is possible that these are monthly recurring orders that get released on a given day of the month.  If that is the case, what are the sales reps doing the rest of the month?
*  The lines with multiple days, unique_days > 1, have a noticeable higher number of orders, so_cnt, and associated so dollars.

The plot clearly shows that two months with multiple sales rep order days for 2011, (1108 and 1110), one for 2012, (1201), and two in 2014, (1401 and 1403).  The 1403 is the only three day sales rep order month.

In the next code block, we flesh out the dates associatd with the sales reps' orders.  Since 4 out of the 5 months with multiple order days only have two dates, the code block captures them with a min/max orderdate.

```{r}
mo_so_mo_dt_dist_sum <- dbGetQuery(con,
"
SELECT EXTRACT(YEAR FROM soh.orderdate) yr
     , EXTRACT(MONTH FROM soh.orderdate) mo
--     , EXTRACT(DOW FROM soh.orderdate) dow
--     , case when onlineorderflag then 'online' else 'sales rep' end so_type
     , COUNT(*) so_cnt
     , sum(subtotal) so_dollars
     , min(soh.orderdate) min_dt
     , max(soh.orderdate) max_dt
     , EXTRACT(DAY FROM max(soh.orderdate)) - EXTRACT(DAY FROM min(soh.orderdate)) + 1 days
     , count(distinct soh.orderdate) unique_days
  FROM sales.salesorderheader soh
 where not onlineorderflag
group by EXTRACT(YEAR FROM orderdate)  
     , EXTRACT(MONTH FROM orderdate)
--     , EXTRACT(DOW FROM soh.orderdate) 
     , case when onlineorderflag then 'online' else 'sales rep' end
order by yr, mo
")
mo_so_mo_dt_dist_sum
# so_rep_dist <- mo_so_mo_dt_dist_sum %>% filter (unique_days > 1)
# so_rep_dist
```

Looking at the min-max dt columns, most sales rep orders are entered on or near the last day of the month and or the first day of the month.

Looking at 2011 we see an entry for May, but not June.  However, July only has an entry 7-1, but not a 7-31 entry.  Perhaps the missing 6-30 entry is actually the 7-1 entry.  If the 7-1 is really the 6-30 entry, then the same pattern would make the 8-1 entry, the real 7-31 entry.  There is no 9-31 entry,  but there is a 10-1 and a 10-31 entry.  The same patterns adjustment has the 10-1 entry for the missing 9-31 entry.  The same adjustment pattern holds for the missing 11-30 entry is the 12-1-29 entries.

The next code block shows the details associated with the sales rep orders with multiple order days.

```{r  sales rep multi-day order months }
mo_so_mo_dt_dist_xdays <- dbGetQuery(con,
"
SELECT to_char(orderdate,'YYMM') yymm
     ,soh.orderdate
     , COUNT(DISTINCT soh.orderdate) *1.0 unique_days
     , COUNT(*) so_cnt
     , sum(subtotal) so_dollars
  FROM sales.salesorderheader soh
where not onlineorderflag
  and to_char(orderdate,'YYMM') in ('1108','1110','1201','1401','1403')
group by to_char(orderdate,'YYMM') 
     , soh.orderdate
ORDER BY soh.orderdate
")
mo_so_mo_dt_dist_xdays
```

# Adjusted Sales Rep Order Dates

In the next code block, the orders entered on the first of the month are adjusted backwards one day to put them in the prior month.

```{r}
x <- dbGetQuery(con,
"
select to_char(orderdate_adj,'YYMM') yymm
      ,YR_adj yr,mo_adj mo
      ,sales_type
      ,sum(so_cnt) so_cnt
      ,sum(so_dollars) so_dollars
  from (
SELECT EXTRACT(YEAR FROM soh.orderdate) yr
     , EXTRACT(MONTH FROM soh.orderdate) mo
     , EXTRACT(DAY FROM soh.orderdate) dd
     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN  soh.orderdate - '1 day'::interval
            ELSE  soh.orderdate
       END AS orderdate_adj
     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN EXTRACT(YEAR FROM soh.orderdate - '1 day'::interval)
            ELSE  EXTRACT(YEAR FROM soh.orderdate) 
       END AS YR_adj

     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN EXTRACT(MONTH FROM soh.orderdate - '1 day'::interval)
            ELSE EXTRACT(MONTH FROM soh.orderdate) 
       END mo_adj
     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN EXTRACT(DAY FROM soh.orderdate - '1 day'::interval)
            ELSE EXTRACT(DAY FROM soh.orderdate) 
       END dd_adj
     , COUNT(*) * 1.0 so_cnt 
     , sum(subtotal) so_dollars
     , min(soh.orderdate) min_dt
     , max(soh.orderdate) max_dt
     , EXTRACT(DAY FROM max(soh.orderdate)) - EXTRACT(DAY FROM min(soh.orderdate)) + 1 days
     , count(distinct soh.orderdate) unique_days
     , case when soh.onlineorderflag then 'online' else 'sales rep' end sales_type
  FROM sales.salesorderheader soh
 where not onlineorderflag
group by EXTRACT(YEAR FROM soh.orderdate)
     , EXTRACT(MONTH FROM soh.orderdate)
     , EXTRACT(DAY FROM soh.orderdate)
     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN  soh.orderdate - '1 day'::interval
            ELSE  soh.orderdate
       END
     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN EXTRACT(YEAR FROM soh.orderdate - '1 day'::interval)
            ELSE  EXTRACT(YEAR FROM soh.orderdate) 
       END
     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN EXTRACT(MONTH FROM soh.orderdate - '1 day'::interval)
            ELSE EXTRACT(MONTH FROM soh.orderdate) 
       END
     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN EXTRACT(DAY FROM soh.orderdate - '1 day'::interval)
            ELSE EXTRACT(DAY FROM soh.orderdate) 
       END
     , case when onlineorderflag then 'online' else 'sales rep' end
) as src
group by yymm,YR_adj ,mo_adj ,sales_type
ORDER BY mo, yr, sales_type
")
x

```

```{r Sales Rep Orders Adjusted PLot, fig.width=10}
x$yymm <- as.factor(x$yymm)
x$mo <- as.factor(x$mo)
x$yr <- as.factor(x$yr)

#  x$sales_type <- as.factor(mo_onl_pct$sales_type)
mo_2011 <- x %>% filter (yr == 2011)
mo_2012 <- x %>% filter (yr == 2012)
mo_2013 <- x %>% filter (yr == 2013)
mo_2014 <- x %>% filter (yr == 2014)

ggplot(data=NULL) +
  # data=mo_2011 first results in the x axis months out of order.
  geom_line (data=mo_2012,aes(x=mo,y=so_cnt,color = yr,group = sales_type)) +
  geom_line (data=mo_2011,aes(x=mo,y=so_cnt,color = yr,group = sales_type)) +
  geom_line (data=mo_2013,aes(x=mo,y=so_cnt,color = yr,group = sales_type)) +
  geom_line (data=mo_2014,aes(x=mo,y=so_cnt,color = yr,group = sales_type)) +
  
  geom_point(data=mo_2011,aes(x=mo,y=so_cnt,color = sales_type)) +  
  geom_point(data=mo_2012,aes(x=mo,y=so_cnt,color = sales_type)) +
  geom_point(data=mo_2013,aes(x=mo,y=so_cnt,color = sales_type)) +
  geom_point(data=mo_2014,aes(x=mo,y=so_cnt,color = sales_type)) +
  
  xlab("Month") + 
  ylab("Sales Rep") +
  theme(plot.title = element_text(hjust = .50)) +  
  ggtitle(paste("Sales Rep Orders by Month"))



```

## Sales Rep Orders By Fiscal Year


```{r}
SalesRepOrders_Fiscal_Year <- dbGetQuery(con,
"
select to_char(orderdate_adj,'YYMM') yymm
      ,fiscalyear_adj
      ,YR_adj yr,mo_adj mo
      ,sales_type
      ,sum(so_cnt) so_cnt
      ,sum(so_dollars) so_dollars
  from (
SELECT date_part('year'::text, soh.orderdate + '6 mons'::interval) AS fiscalyear
     , EXTRACT(YEAR FROM soh.orderdate) yr
     , EXTRACT(MONTH FROM soh.orderdate) mo
     , EXTRACT(DAY FROM soh.orderdate) dd
     , date_part('year'::text,case when EXTRACT(DAY FROM soh.orderdate) = 1
                                   THEN  soh.orderdate - '1 day'::interval
                                   ELSE  soh.orderdate
                              END+ '6 months'::interval) AS fiscalyear_adj 
     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN  soh.orderdate - '1 day'::interval
            ELSE  soh.orderdate
       END AS orderdate_adj
     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN EXTRACT(YEAR FROM soh.orderdate - '1 day'::interval)
            ELSE  EXTRACT(YEAR FROM soh.orderdate) 
       END AS YR_adj

     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN EXTRACT(MONTH FROM soh.orderdate - '1 day'::interval)
            ELSE EXTRACT(MONTH FROM soh.orderdate) 
       END mo_adj
     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN EXTRACT(DAY FROM soh.orderdate - '1 day'::interval)
            ELSE EXTRACT(DAY FROM soh.orderdate) 
       END dd_adj
     , COUNT(*) * 1.0 so_cnt 
     , ROUND(sum(subtotal),2) so_dollars
     , min(soh.orderdate) min_dt
     , max(soh.orderdate) max_dt
     , EXTRACT(DAY FROM max(soh.orderdate)) - EXTRACT(DAY FROM min(soh.orderdate)) + 1 days
     , count(distinct soh.orderdate) unique_days
     , case when soh.onlineorderflag then 'online' else 'sales rep' end sales_type
  FROM sales.salesorderheader soh
 where not onlineorderflag
group by date_part('year'::text, soh.orderdate + '6 mons'::interval) 
     , EXTRACT(YEAR FROM soh.orderdate)
     , EXTRACT(MONTH FROM soh.orderdate)
     , EXTRACT(DAY FROM soh.orderdate)
     , date_part('year'::text,case when EXTRACT(DAY FROM soh.orderdate) = 1
                                   THEN  soh.orderdate - '1 day'::interval
                                   ELSE  soh.orderdate
                              END+ '6 months'::interval)
     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN  soh.orderdate - '1 day'::interval
            ELSE  soh.orderdate
       END
     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN EXTRACT(YEAR FROM soh.orderdate - '1 day'::interval)
            ELSE  EXTRACT(YEAR FROM soh.orderdate) 
       END
     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN EXTRACT(MONTH FROM soh.orderdate - '1 day'::interval)
            ELSE EXTRACT(MONTH FROM soh.orderdate) 
       END
     , case when EXTRACT(DAY FROM soh.orderdate) = 1
            THEN EXTRACT(DAY FROM soh.orderdate - '1 day'::interval)
            ELSE EXTRACT(DAY FROM soh.orderdate) 
       END
     , case when onlineorderflag then 'online' else 'sales rep' end
) as src
group by yymm,fiscalyear_adj,YR_adj ,mo_adj ,sales_type
ORDER BY case when mo_adj > 6 then mo_adj - 6 else mo_adj + 6 end,fiscalyear_adj
--ORDER BY fiscalyear_adj,case when mo_adj > 6 then mo_adj - 6 else mo_adj + 6 end
--ORDER BY mo, yr, sales_type
")

```


```{r fig.width=16}
SalesRepOrders_Fiscal_Year$mo <- factor(SalesRepOrders_Fiscal_Year$mo,levels=c('7','8','9','10','11','12','1','2','3','4','5','6'))

ggplot(data=SalesRepOrders_Fiscal_Year
       ,aes(x=mo,y=so_cnt,color=as.factor(fiscalyear_adj),group = as.factor(fiscalyear_adj))) + 
  geom_line() +
  geom_point() +
  geom_text(aes(x=mo,y=so_cnt,label=so_dollars)
           ,color = 'black'
           ,vjust = -1.5) +
  labs(color='Fiscal Year') +
  xlab("Month") + 
  ylab("Orders") +
  theme(plot.title = element_text(hjust = .50)) +  
  ggtitle(paste("Sales Rep Orders by Month\nUsing Adjusted Order Dates "))
```

```{r}
mo_so_mo_dt_dist_det <- dbGetQuery(con,
"
SELECT EXTRACT(MONTH FROM soh.orderdate) mo
     , EXTRACT(YEAR FROM soh.orderdate) yr
     , soh.orderdate
     , case when onlineorderflag then 'online' else 'sales rep' end so_type
     , sum(subtotal)/10000 so_dollars
     , COUNT(*) * 1.0 so_cnt
  FROM sales.salesorderheader soh
group by EXTRACT(MONTH FROM orderdate) 
     , EXTRACT(YEAR FROM orderdate) 
     , case when onlineorderflag then 'online' else 'sales rep' end
     ,soh.orderdate
order by case when onlineorderflag then 'online' else 'sales rep' end desc,soh.orderdate
")
long <- gather(data=mo_so_mo_dt_dist_det,key='sum_type','sum_amt',so_cnt,so_dollars)
sp_print_df(long)
long_online <- long %>% filter(so_type == 'sales rep')

ggplot(data = long_online, aes(x=orderdate,y=sum_amt,color=sum_type)) +
  geom_line() 

```

```{r missing sales rep order dates}
dbGetQuery(con,
"
with all_yymm as (
  select distinct to_char(orderdate,'YYMM') yymm

  FROM sales.salesorderheader soh
)
, salesrep_orders as (
  SELECT distinct to_char(orderdate,'YYMM') yymm
  FROM sales.salesorderheader soh
  where not onlineorderflag
)
select * 
  from all_yymm left outer join salesrep_orders on all_yymm.yymm = salesrep_orders.yymm
 where salesrep_orders.yymm is null
 order by all_yymm.yymm desc
")
```


