---
title: "AW SO EDA"
author: "Sophie Yang"
date: "July 19, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=10)
```

```{r libraries, include=FALSE}
library(tidyverse)
library(DBI)
library(RPostgres)
library(glue)
require(knitr)
library(dbplyr)
library(sqlpetr)
library(bookdown)
library(here)

library(scales)  # ggplot xy scales
```

```{r, start adventureworks}
sp_docker_start('adventureworks')
# Sys.sleep(10)
con <- sp_get_postgres_connection(
  host = "localhost",
  port = 5432,
  user = "postgres",
  password = "postgres",
  dbname = "adventureworks",
  seconds_to_test = 60, connection_tab = TRUE
)
```

```{r so_adj_date function}
#
# This code block creates a SQL function to correct the sales order dates for sales rep orders.  If the passed order date
# is the 1'st, the sales order date is adjusted back one day.  No adjustment is made for on-line orders

# adventureworks schema vs public schma?  
# assumes root privileges for adventureworks

#
dbExecute(con,
"CREATE OR REPLACE FUNCTION so_adj_date(so_date timestamp, ONLINE_ORDER boolean) RETURNS timestamp AS $$
     BEGIN
        IF (ONLINE_ORDER) THEN
            RETURN (SELECT so_date);
        ELSE
            RETURN(SELECT CASE WHEN EXTRACT(DAY FROM so_date) = 1
                               THEN  so_date - '1 day'::interval
                               ELSE  so_date
                          END
                  );
        END IF;
 END; $$
LANGUAGE PLPGSQL;
")
```

```{r Dry Version so_adj_date function}
SalesRepOrders_Fiscal_Year <- dbGetQuery(con,
"
with adj_so_dates as
       (SELECT date_part('year'::text, soh.orderdate + '6 mons'::interval) AS fiscalyear
             , soh.orderdate
             , EXTRACT(YEAR FROM soh.orderdate) yr
             , EXTRACT(MONTH FROM soh.orderdate) mo
             , EXTRACT(DAY FROM soh.orderdate) dd
             , date_part('year'::text,so_adj_date(soh.orderdate, onlineorderflag) + '6 months'::interval) AS fiscalyear_adj
             , so_adj_date(soh.orderdate, onlineorderflag) orderdate_adj
             , EXTRACT(YEAR FROM so_adj_date(soh.orderdate, onlineorderflag)) YR_adj
             , EXTRACT(MONTH FROM so_adj_date(soh.orderdate, onlineorderflag)) mo_adj
             , EXTRACT(DAY FROM so_adj_date(soh.orderdate, onlineorderflag)) dd_adj
             , COUNT(*) * 1.0 so_cnt 
             , ROUND(sum(subtotal),2) so_dollars
             , min(soh.orderdate) min_dt
             , max(soh.orderdate) max_dt
             , EXTRACT(DAY FROM max(soh.orderdate)) - EXTRACT(DAY FROM min(soh.orderdate)) + 1 days
             , count(distinct soh.orderdate) unique_days
             , case when soh.onlineorderflag then 'online' else 'sales rep' end sales_type
          FROM sales.salesorderheader soh
         where not onlineorderflag
        group by date_part('year'::text, soh.orderdate + '6 mons'::interval) 
             , soh.orderdate
             , EXTRACT(YEAR FROM soh.orderdate)
             , EXTRACT(MONTH FROM soh.orderdate)
             , EXTRACT(DAY FROM soh.orderdate)
             , case when onlineorderflag then 'online' else 'sales rep' end
             , so_adj_date(soh.orderdate, onlineorderflag)
             , date_part('year'::text,so_adj_date(soh.orderdate, onlineorderflag) + '6 months'::interval) 
             , EXTRACT(YEAR FROM so_adj_date(soh.orderdate, onlineorderflag))
             , EXTRACT(MONTH FROM so_adj_date(soh.orderdate, onlineorderflag))
             , EXTRACT(DAY FROM so_adj_date(soh.orderdate, onlineorderflag))
        ) 
--select * from adj_so_dates
select to_char(orderdate_adj,'YYMM') yymm
      ,fiscalyear_adj,YR_adj yr,mo_adj mo
      ,sales_type
      ,sum(so_cnt) so_cnt
      ,sum(so_dollars) so_dollars
  from adj_so_dates
group by yymm,fiscalyear_adj,YR_adj ,mo_adj ,sales_type
ORDER BY case when mo_adj > 6 then mo_adj - 6 else mo_adj + 6 end,fiscalyear_adj
")
sp_print_df(SalesRepOrders_Fiscal_Year)
```


```{r fig.width=16}
SalesRepOrders_Fiscal_Year$mo <- factor(SalesRepOrders_Fiscal_Year$mo,levels=c('7','8','9','10','11','12','1','2','3','4','5','6'))

ggplot(data=SalesRepOrders_Fiscal_Year
       ,aes(x=mo,y=so_cnt,color=as.factor(fiscalyear_adj),group = as.factor(fiscalyear_adj))) + 
  geom_line() +
  geom_point() +
  geom_text(aes(x=mo,y=so_cnt,label=so_dollars)
           ,color = 'black'
           ,vjust = -1.5) +
  labs(color='Fiscal Year') +
  xlab("Month") + 
  ylab("Orders") +
  theme(plot.title = element_text(hjust = .50)) +  
  ggtitle(paste("Sales Rep Orders by Month\nUsing Adjusted Order Dates "))
```


```{r}
product_count <- dbGetQuery(con,"select makeflag,finishedgoodsflag,discontinueddate
                                       ,max(daystomanufacture) daysToMfg,count(*) 
                                   from production.product p 
                                   group by makeflag,finishedgoodsflag,discontinueddate")
product_count
```

```{r How many sellable items have never sold}
x <- dbGetQuery(con,
"select makeflag,finishedgoodsflag
       ,max(daystomanufacture) daysToMfg,count(distinct p.productid) prod_cnt,count(distinct sod.productid) sold_prod_cnt
       ,count(distinct p.productid) -count(distinct sod.productid) never_sold_prod_cnt
   from production.product p left join sales.salesorderdetail sod on p.productid = sod.productid
  group by makeflag,finishedgoodsflag
  order by finishedgoodsflag
")
x
```

```{r What sellable items have never sold}
x <- dbGetQuery(con,
"select makeflag,p.productid,name,standardcost,sellstartdate,sellenddate
       ,sum(coalesce(pi.quantity,0)) pi_qty
   from production.product p 
        left join production.productinventory pi on p.productid = pi.productid
        left join sales.salesorderdetail sod on p.productid = sod.productid
  where finishedgoodsflag and sod.productid is null
 group by makeflag,p.productid,name,standardcost,sellstartdate,sellenddate
")
x
```

```{r What sellable items have never sold exist in inventory}
x <- dbGetQuery(con,
"select makeflag,p.productid,name,standardcost
       ,sum(coalesce(pi.quantity,0)) qty
       ,sum(coalesce(pi.quantity,0)*standardcost) extcost
       ,pi.modifieddate,sellstartdate,sellenddate
   from production.product p 
        left join production.productinventory pi on p.productid = pi.productid
        left join sales.salesorderdetail sod on p.productid = sod.productid
  where finishedgoodsflag and sod.productid is null
    and coalesce(pi.quantity,0) > 0
 group by makeflag,p.productid,name,standardcost,pi.modifieddate,sellstartdate,sellenddate
")
x
```
```



```{r}
top10_products_by_month <- dbGetQuery(con,
"
With product_details as
(
        select so_adj_date(soh.orderdate, onlineorderflag) orderdate_adj
              ,date_part('year'::text,so_adj_date(soh.orderdate, onlineorderflag)+ '6 months'::interval) AS fiscalyear_adj 
              ,to_char(so_adj_date(soh.orderdate, onlineorderflag),'YYMM') AS YYMM
              ,sod.productid,name
              ,orderqty
              ,unitprice
              ,orderqty * unitprice ext_price
              ,onlineorderflag
          from sales.salesorderheader soh,sales.salesorderdetail sod
              ,production.product p
         where soh.salesorderid = sod.salesorderid
           and sod.productid = p.productid 
--and sod.productid between 740 and 750
and to_char(so_adj_date(soh.orderdate, onlineorderflag),'YYMM') = '1105'
)
--select * from product_details limit 100 
/*
select yymm,productid,sum(orderqty),unitprice,sum(ext_price),onlineorderflag from product_details
group by yymm,productid,unitprice,onlineorderflag
order by yymm
*/

select yymm,productid,name
      ,sum(orderqty) orderqty
      ,sum(ext_price) ext_price
      ,sum(case when onlineorderflag then orderqty else 0 end) online_orders
      ,sum(case when onlineorderflag then ext_price else 0 end) olo_ext_price
      ,sum(case when not onlineorderflag then orderqty else 0 end) sales_rep_orders
      ,sum(case when not onlineorderflag then ext_price else 0 end) sro_ext_price
  from product_details
group by yymm,productid,name
order by yymm
"
)
sp_print_df(top10_products_by_month)
```


```{r}
top10_products_by_month <- dbGetQuery(con,
"
select orderdate,fiscalyear_adj,yymm
      ,productid,name
      ,orderqty
      ,extprice 
      --,rn
  from (
        select soh.orderdate
              ,date_part('year'::text,so_adj_date(soh.orderdate, onlineorderflag)+ '6 months'::interval) AS fiscalyear_adj 
              ,to_char(so_adj_date(soh.orderdate, onlineorderflag),'YYMM') AS YYMM
              ,sod.productid,name
              ,sum(orderqty) orderqty,sum(unitprice*orderqty) extprice 
 --             ,ROW_NUMBER () OVER (PARTITION BY to_char(so_adj_date(soh.orderdate, onlineorderflag),'YYDD'), sod.productid
  --                                 ORDER BY sum(unitprice*orderqty) desc) rn
          from sales.salesorderheader soh,sales.salesorderdetail sod
             , production.product p
         where soh.salesorderid = sod.salesorderid
           and sod.productid = p.productid 
and sod.productid = 749
        group by soh.orderdate
                ,date_part('year'::text,so_adj_date(soh.orderdate, onlineorderflag)+ '6 months'::interval)
                ,to_char(so_adj_date(soh.orderdate, onlineorderflag),'YYMM')
                ,sod.productid,name
        ) as src
 --where rn <= 10
order by yymm, productid    
")
sp_print_df(top10_products_by_month)
```




```{r}
x <- top10_products_by_month %>% filter(yymm %in% c('1105'))
ggplot(data=x,aes(x=as.factor(productid),y=extprice))+
  geom_col()  

ggplot(data=x,aes(x=extprice,y=reorder(productid,extprice)) ) +
  geom_point() +
  theme_light() +
  theme(panel.grid.major.x = element_blank()
       ,panel.grid.minor.x = element_blank()
       ,panel.grid.minor.y = element_line(color='grey60',linetype = 'dashed')
       ) +
  xlab('Extended Price') +
  ylab('Product ID') 
```






```{r top 10 Sales Rep Products By Month}
x <- dbGetQuery(con,
"
select yymm
      ,productid,name
      ,orderqty
      ,extprice 
      ,rn
  from (
select date_part('year'::text,case when EXTRACT(DAY FROM soh.orderdate) = 1
                                   THEN  soh.orderdate - '1 day'::interval
                                   ELSE  soh.orderdate
                              END+ '6 months'::interval) AS fiscalyear_adj 
      ,to_char(case when EXTRACT(DAY FROM soh.orderdate) = 1 THEN  soh.orderdate - '1 day'::interval ELSE soh.orderdate END,'YYMM') AS YYMM
      ,sod.productid,p.name
      ,unitprice
      ,sum(orderqty) orderqty,sum(unitprice*orderqty) extprice 
      ,ROW_NUMBER () OVER (PARTITION BY to_char(case when EXTRACT(DAY FROM soh.orderdate) = 1 
                                                     THEN soh.orderdate - '1 day'::interval 
                                                     ELSE soh.orderdate 
                                                END,'YYMM')
                           ORDER BY sum(unitprice*orderqty) desc) rn
  from sales.salesorderheader soh,sales.salesorderdetail sod
     , production.product p
--     , production.vproductanddescription p
 where soh.salesorderid = sod.salesorderid
   and sod.productid = p.productid
   and not onlineorderflag
group by date_part('year'::text,case when EXTRACT(DAY FROM soh.orderdate) = 1
                                   THEN  soh.orderdate - '1 day'::interval
                                   ELSE  soh.orderdate
                              END+ '6 months'::interval)
        ,to_char(case when EXTRACT(DAY FROM soh.orderdate) = 1 THEN  soh.orderdate - '1 day'::interval ELSE soh.orderdate END,'YYMM')
        ,sod.productid,p.name
        ,unitprice
) as src
where rn <= 10
order by yymm, extprice desc 
")


```
How many products are sold

```{r}
dbGetQuery(con,
"select p.*
--  from production.productmodel p
  from production.vproductanddescription p

")
```


```{r}
prod_sales_onl_sr <- dbGetQuery(con,
"select p.productid,p.name
       ,sum(case when onlineorderflag then 1 else 0 end) onl
       ,sum(case when not onlineorderflag then 1 else 0 end) salesrep
       ,count(*) cnt
  from sales.salesorderheader soh,sales.salesorderdetail sod, production.vproductanddescription p
 where soh.salesorderid = sod.salesorderid
   and sod.productid = p.productid
group by p.productid,p.name
")

```

Which items have only sold on line

```{r}
prod_sales_onl_sr %>% 
  filter(onl > 0 & salesrep == 0)
```

Which items have only been sold via a sales rep

```{r}
prod_sales_onl_sr %>% 
  filter(onl == 0 & salesrep > 0)
```

Which items have been sold both online and via a sales rep

```{r}
prod_sales_onl_sr %>% 
  filter(onl > 0 & salesrep > 0)
```


```{r}
dbGetQuery(con,
"with x as
(select soh.salesorderid,soh.subtotal
       ,sum(sod.orderqty*sod.unitprice) det_subtotal1
       ,sum(sod.orderqty*(sod.unitprice-(1.0-sod.unitpricediscount))) det_subtotal2
       ,sum(sod.orderqty*(sod.unitprice-sod.unitpricediscount)) det_subtotal3
       ,count(*) lines
  from sales.salesorderheader soh,sales.salesorderdetail sod
     , production.product p
--     , production.vproductanddescription p
 where soh.salesorderid = sod.salesorderid
   and sod.productid = p.productid 
--   and sod.salesorderid = 43659
group by soh.salesorderid,soh.subtotal)

select * from x where subtotal <> det_subtotal2
")
```

```{r}
dbGetQuery(con,
"with x as
(select soh.salesorderid,soh.subtotal,sod.orderqty,sod.unitprice,sod.unitpricediscount
,sod.orderqty*(sod.unitprice) ext_subtot
,sod.orderqty*(sod.unitprice* (1.0-sod.unitpricediscount)) ext_subtot_Pct
,sod.orderqty*(sod.unitprice-sod.unitpricediscount) ext_subtot_dsc

,sod.specialofferid
  from sales.salesorderheader soh,sales.salesorderdetail sod
     , production.product p
 where soh.salesorderid = sod.salesorderid
   and sod.productid = p.productid 
   and sod.salesorderid = 44098
)
select * from x 
")

```





```{r}
x1 <- dbGetQuery(con,
"
select date_part('year'::text,case when EXTRACT(DAY FROM soh.orderdate) = 1
                                   THEN  soh.orderdate - '1 day'::interval
                                   ELSE  soh.orderdate
                              END+ '6 months'::interval) AS fiscalyear_adj 
      ,to_char(case when EXTRACT(DAY FROM soh.orderdate) = 1 THEN  soh.orderdate - '1 day'::interval ELSE soh.orderdate END,'YYDD') AS YYMM
      ,sod.productid,name
      ,unitprice
      ,sum(orderqty) orderqty,sum(unitprice*orderqty) extprice 
      ,ROW_NUMBER () OVER (PARTITION BY to_char(case when EXTRACT(DAY FROM soh.orderdate) = 1 
                                                     THEN soh.orderdate - '1 day'::interval 
                                                     ELSE soh.orderdate 
                                                END,'YYDD')
                           ORDER BY sum(unitprice*orderqty) desc) rn
  from sales.salesorderheader soh,sales.salesorderdetail sod, production.vproductanddescription p
 where soh.salesorderid = sod.salesorderid
   and sod.productid = p.productid 
group by date_part('year'::text,case when EXTRACT(DAY FROM soh.orderdate) = 1
                                   THEN  soh.orderdate - '1 day'::interval
                                   ELSE  soh.orderdate
                              END+ '6 months'::interval)
        ,to_char(case when EXTRACT(DAY FROM soh.orderdate) = 1 THEN  soh.orderdate - '1 day'::interval ELSE soh.orderdate END,'YYDD')
        ,sod.productid,name
        ,unitprice
")
```





```{r}
head(dbGetQuery(con,
"
select orderdate,onlineorderflag,so_adj_date(orderdate,onlineorderflag),count(*),sum(subtotal)
  from sales.salesorderheader 
 where EXTRACT(DAY FROM orderdate) = 1
  and not onlineorderflag
group by orderdate,onlineorderflag,so_adj_date(orderdate,onlineorderflag)
"))
```

```{r}
dbGetQuery(con,
"select n.nspname as function_schema,
       p.proname as function_name,
       l.lanname as function_language,
       case when l.lanname = 'internal' then p.prosrc
            else pg_get_functiondef(p.oid)
            end as definition,
       pg_get_function_arguments(p.oid) as function_arguments,
       t.typname as return_type
from pg_proc p
left join pg_namespace n on p.pronamespace = n.oid
left join pg_language l on p.prolang = l.oid
left join pg_type t on t.oid = p.prorettype 
where n.nspname not in ('pg_catalog', 'information_schema')
  and p.proname = 'so_adj_date'
order by function_schema,
         function_name;
")
```
```{r drop previous date function attempts}
# dbGetQuery(con,'drop function public.so_adj_date(so_date date)')

# dbGetQuery(con,'drop function  public.so_adj_date(so_date timestamp without time zone)')
```


