---
title: "dbReadTable not lazy Versus tbl lazy"
author: "Sophie Yang"
date: "March 8, 2019"
output: html_document
---



```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# These packages are called in almost every chapter of the book:
library(tidyverse)
library(DBI)
library(RPostgres)
library(glue)
library(here)
require(knitr)
library(dbplyr)
library(sqlpetr)
```

Verify Docker is up and running:

```{r}
sp_check_that_docker_is_up()
```

Verify pet DB is available, it may be stopped.

```{r}
sp_show_all_docker_containers()
```

Start up the `docker-pet` container

```{r}
sp_docker_start("sql-pet")
```

Now connect to the database with R

```{r}

# need to wait for Docker & Postgres to come up before connecting.

con <- sp_get_postgres_connection(
  user = Sys.getenv("DEFAULT_POSTGRES_USER_NAME"),
  password = Sys.getenv("DEFAULT_POSTGRES_PASSWORD"),
  dbname = "dvdrental",
  seconds_to_test = 30
)
```



|Local Processing                   | Remote Processing
|-----------------------------------|---------------------------------------
|Desktops/laptops have fewer resources than the database server.|computer under powered compared to database server|Databaseserver are high powered machines.
|Query result must fit into memory. |Servers have a lot of memory
|Data must be pulled over the network.|Data moved locally over the server backbone.
|Result set can be saved locally.   |With proper privileges one can save results in the database
|R packages availabe                |May require creating R pseudo functions to call a database functions
|

|Operation    |dplyr<br>Local-Only|dplyr<br>Local-Lazy|SQL
|-------------|-------------------|-------------------|----------------
|connection|DBI::dbConnect|DBI::dbConnect|DBI::dbConnect
|Single Read joining one or more tables<br>and fits into memory|DBI::dbReadTable, (returns a df)<br>R package calls always available|tbl() + collect()<br>tbl:Returns two lists<br> collect(): returns tbl_df<br>R package calls available only after collect() call.  Ideally push everything to DB.<br>May require R placeholder functions to mimick DB functions.|dbGetQuery
|Multiple Reads|Not Applicable|Not Applicable|dbSendQuery + dbFetch + dbClearResult
|Fetch Data locally|DBI::dbReadTable fetches data|collect()|dbGetQuery or dbSendQuery+dbFetch+dbClearResult|dbGetQuery or dbSendQuery + dbFetch
|Write Results Local|write family of functions|write family of functions|write family of functions
|Write Results to DB|compute() or copy_to|compute() or copy_to|compute() or copy_to

                     
## Dplyr Local-Only

```{r}
rental_table <- tbl(con,'rental')
inventory_table <- tbl(con,'inventory')
film_table <- tbl(con,'film')
rental_table %>% collect(n=10) %>% mutate(YYYY_MM = format(rental_date,"%Y-%m")) %>% select(rental_date,YYYY_MM)
```

Without the collect(), the processing is performed remotely on the database server and the `format(rental_date,"%Y-%m")` fails with the following error message.

<pre>
Error in result_create(conn@ptr, statement) : 
  Failed to prepare query: ERROR:  function format(timestamp without time zone, unknown) does not exist
LINE 1: ...", "last_update", "special_features", "fulltext", FORMAT("re...
                                                             ^
HINT:  No function matches the given name and argument types. You might need to add explicit type casts.
</pre>

## Dplyr Local-Lazy Approach

<pre>
to_char</b> <- function(dt, fmt) {return(fmt)}

rental_table %>% inner_join(inventory_table, by = c("inventory_id" = "inventory_id"), suffix(c(".r", ".i"))) %>% inner_join(film_table, by = c("film_id" = "film_id"), suffix(c(".i", ".f"))) %>% mutate(YYYY_MM = to_char(rental_date,"YYYY-MM"))
</pre>

```{r}
## to_char <- function(dt, fmt) {return(fmt)}
rental_table <- tbl(con,'rental')
rental_table  %>% mutate(YYYY_MM = to_char(rental_date,"YYYY-MM"))%>%select(rental_date,YYYY_MM)
```

Now I cannot get the error message I thought I got in the past and got in today's phone call.  The function has been commented out.
 
When the code blocks are executed one at a time, the above code block returns 1000 rows, not LAZY.   However, when knitted, it is LAZY.

```{r}
#to_char <- function(dt, fmt) {return(fmt)}
rental_table <- tbl(con,'rental')
inventory_table <- tbl(con,'inventory')
film_table <- tbl(con,'film')
rental_table %>% inner_join(inventory_table, by = c("inventory_id" = "inventory_id"), suffix(c(".r", ".i"))) %>% inner_join(film_table, by = c("film_id" = "film_id"), suffix(c(".i", ".f"))) %>% mutate(YYYY_MM = to_char(rental_date,"YYYY-MM")) %>% select(rental_date,YYYY_MM)
```
 
Without the declared `to_char` function, the above code block returns the following error.

<pre>
Error in eval(lhs, parent, parent) : object 'rental_table' not found
</pre>