---
title: "dbReadTable not lazy Versus tbl lazy"
author: "Sophie Yang"
date: "March 8, 2019"
output: html_document
---



```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# These packages are called in almost every chapter of the book:
library(tidyverse)
library(DBI)
library(RPostgres)
library(glue)
library(here)
require(knitr)
library(dbplyr)
library(sqlpetr)
```

Verify Docker is up and running:

```{r}
sp_check_that_docker_is_up()
```

Verify pet DB is available, it may be stopped.

```{r}
sp_show_all_docker_containers()
```

Start up the `docker-pet` container

```{r}
sp_docker_start("sql-pet")
```

Now connect to the database with R

```{r}

# need to wait for Docker & Postgres to come up before connecting.

con <- sp_get_postgres_connection(
  user = Sys.getenv("DEFAULT_POSTGRES_USER_NAME"),
  password = Sys.getenv("DEFAULT_POSTGRES_PASSWORD"),
  dbname = "dvdrental",
  seconds_to_test = 30
)
```



|Local Processing                   | Remote Processing
|-----------------------------------|---------------------------------------
|Desktops/laptops have fewer resources than the database server.|computer under powered compared to database server|Databaseserver are high powered machines.
|Query result must fit into memory. |Servers have a lot of memory
|Data must be pulled over the network.|Data moved locally over the server backbone.
|Result set can be saved locally.   |With proper privileges one can save results in the database
|R packages availabe                |May require creating R pseudo functions to call a database functions
|

|Operation    |dplyr<br>Local-Only|dplyr<br>Local-Lazy|SQL
|-------------|-------------------|-------------------|----------------
|connection|DBI::dbConnect|DBI::dbConnect|DBI::dbConnect
|Single Read joining one or more tables<br>and fits into memory|DBI::dbReadTable, (returns a df)<br>R package calls always available|tbl() + collect()<br>tbl:Returns two lists<br> collect(): returns tbl_df<br>R package calls available only after collect() call.  Ideally push everything to DB.<br>May require R placeholder functions to mimick DB functions.|dbGetQuery
|Multiple Reads|Not Applicable|Not Applicable|dbSendQuery + dbFetch + dbClearResult
|Fetch Data locally|DBI::dbReadTable fetches data|collect()|dbGetQuery or dbSendQuery+dbFetch+dbClearResult|dbGetQuery or dbSendQuery + dbFetch
|Write Results Local|write family of functions|write family of functions|write family of functions
|Write Results to DB|compute() or copy_to|compute() or copy_to|compute() or copy_to

                     
## Dplyr Local-Only

```{r}
rental_table <- tbl(con,'rental')
inventory_table <- tbl(con,'inventory')
film_table <- tbl(con,'film')
rental_table %>% collect(n=10) %>% mutate(YYYY_MM = format(rental_date,"%Y-%m")) %>% select(rental_date,YYYY_MM)
```

Without the collect(), the processing is performed remotely on the database server and the `format(rental_date,"%Y-%m")` fails with the following error message.

<pre>
Error in result_create(conn@ptr, statement) : 
  Failed to prepare query: ERROR:  function format(timestamp without time zone, unknown) does not exist
LINE 1: ...", "last_update", "special_features", "fulltext", FORMAT("re...
                                                             ^
HINT:  No function matches the given name and argument types. You might need to add explicit type casts.
</pre>

## Dplyr Local-Lazy Approach

```{r}
rental_table <- tbl(con,'rental')
# this returns 1000 rows
rs1000 <- rental_table  %>% mutate(YYYY_MM = to_char(rental_date,"YYYY-MM"))%>%select(rental_date,YYYY_MM) 
rs1000
```

When the code blocks are executed one at a time, the above code block returns 1000 rows.  Based on the output, it is unclear that the result set is lazy or not.  When knitted, it is LAZY and only returns 10 rows.  When collect() is added, the block returns 16045 rows.

Note that the following comes from the knit output of the code block above and the first line explicitly tells us that the output is lazy.

<pre>
## # Source:   lazy query [?? x 2]
## # Database: postgres [postgres@localhost:5432/dvdrental]
##    rental_date         YYYY_MM
##    <dttm>              <chr> 
##  1 2005-05-24 22:54:33 2005-05
##  2 2005-05-24 23:03:39 2005-05
##  3 2005-05-24 23:04:41 2005-05
##  4 2005-05-24 23:05:21 2005-05
##  5 2005-05-24 23:08:07 2005-05
##  6 2005-05-24 23:11:53 2005-05
##  7 2005-05-24 23:31:46 2005-05
##  8 2005-05-25 00:00:40 2005-05
##  9 2005-05-25 00:02:21 2005-05
## 10 2005-05-25 00:09:02 2005-05
## # ... with more rows
</pre>



|function type|dplyr function |Postgres function
|-------------|---------------|------------------
|string       |paste()        |concat()
|string       |substr()       |substr()
|string       |strsplit()     |split_part()
|string       |toupper()      |upper()/initcap()
|string       |tolower()      |lower()
|string       |toTitleCase    |initcap()
