# Miniconda Integration

## Why do this?

A number of R deep learning packages use Python under the hood. RStudio's **keras** [@R-keras] package, for example, works this way. Also, the R **docker** [@R-docker] package works by calling a Python Docker API library from R via **reticulate** [@R-reticulate]. And, of course, you'll probably end up receiving a Jupyter notebook or two even if you're a die-hard RStudio user.

[`Miniconda`](https://conda.io/miniconda.html) is a bare-bones minimalist version of the rather large Anaconda environment. If you're doing Python data science, you probably have the full Anaconda installed already. But for R programmers, we only want enough Python for the R packages that use Python libraries to work. So ... here we go!

## Install the `installr` package.
There's an R package called **installr** [@R-installr] that can run a Windows installer.
```{r eval=FALSE}
if (!require(installr)) install.packages("installr")
library(installr)
```

## Install `Miniconda3`
The following R code chunk will install `Miniconda3.

```{r eval=FALSE}
install.URL("https://repo.continuum.io/miniconda/Miniconda3-latest-Windows-x86_64.exe")
```

Here are the screenshots you'll see:

```{r echo=FALSE}
knitr::include_graphics("screenshots/2018-08-31 16_12_57-Miniconda3 4.5.4 (64-bit) Setup.png")
```
<br>Click `Next`.

***

```{r echo=FALSE}
knitr::include_graphics("screenshots/2018-08-31 16_13_57-Miniconda3 4.5.4 (64-bit) Setup.png")
```
<br>Click `I Agree`.

***

```{r echo=FALSE}
knitr::include_graphics("screenshots/2018-08-31 16_14_23-Miniconda3 4.5.4 (64-bit) Setup.png")
```
<br>`Just Me`, `Next`.

***

```{r echo=FALSE}
knitr::include_graphics("screenshots/2018-08-31 16_16_11-Miniconda3 4.5.4 (64-bit) Setup.png")
```
<br>Choose the install location. The default is your home directory, which on my laptop is a small SSD. So I changed it to the `D` drive, which is a terabyte spinning disk. After you've set the install location, click `Next`.

***

```{r echo=FALSE}
knitr::include_graphics("screenshots/2018-08-31 16_18_47-Miniconda3 4.5.4 (64-bit) Setup.png")
```
<br>Clear both check boxes and click `Install`.

***

```{r echo=FALSE}
knitr::include_graphics("screenshots/2018-08-31 16_21_12-Miniconda3 4.5.4 (64-bit) Setup.png")
```
<br>Click `Next`.

***

```{r echo=FALSE}
knitr::include_graphics("screenshots/2018-08-31 16_21_44-Miniconda3 4.5.4 (64-bit) Setup.png")
```
<br>Clear the check boxes and click `Finish`.

## Install `reticulate`
```{r eval=FALSE}
if (!require(reticulate)) install.packages("reticulate")
library(reticulate)
```
<br>Did it work?
```{r eval=FALSE}
py_discover_config()
```

