# Docker, Postgres, and R

We always load the tidyverse and some other packages, but don't show it unless we are using packages other than `tidyverse`, `DBI`, and `RPostgres`.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(DBI)
library(RPostgres)
```
## Verify that Docker running

Docker commands can be run from a terminal (e.g., the Rstudio Terminal pane) or with a `system()` command.  In this tutorial, we use `system2()` so that all the output that is created externally is shown.  Note that `system2` calls are divided into several parts:

1. The program that you are sending a command to.
2. The parameters or commands that are being sent
3. `stdout = TRUE, stderr = TRUE` are two parameters that are standard in this book, so that the comand's full output is shown in the book.

The `docker version` command returns the details about the docker daemon that is running on your computer.
```{r}
system2("docker", "version", stdout = TRUE, stderr = TRUE)
```

The convention we use in this book is to assemble a command with `paste0` so that the parts of the command can be specified separately.
```{r, eval=TRUE, echo=TRUE}
docker_cmd <- paste0(
  "run -d --name temporary-postgres --publish 5432:5432 ",
  " postgres:10"
)
docker_cmd
```

Submit the command constructed above:
```{r, eval=TRUE, echo=TRUE}
system2("docker", docker_cmd, stdout = TRUE, stderr = TRUE)
```
Docker returns a long string of numbers.  If you are running this command for the first time, Docker is downloading the Postgres image and it takes a bit of time.

The following comand shows that `postgres:10` is still running:
```{r, eval=TRUE, echo=TRUE}
system2("docker", "ps", stdout = TRUE, stderr = TRUE)

```

Create a connection to Postgres after waiting 3 seconds so that Docker has time to do its thing.
```{r}
Sys.sleep(3)

con <- DBI::dbConnect(RPostgres::Postgres(),
                      host = "localhost",
                      port = "5432",
                      user = "postgres",
                      password = "postgres")
```

Show that you can connect but that Postgres database doesn't contain any tables:

```{r}
dbListTables(con)
```

Write `mtcars` to Postgres
```{r}
dbWriteTable(con, "mtcars", mtcars)
```

List the tables in the Postgres database to show that `mtcars` is now there:

```{r}
dbListTables(con)

# list the fields in mtcars:
dbListFields(con, "mtcars")
```

Download the table from the DBMS to a local data frame:
```{r}
mtcars_df <- tbl(con, "mtcars")

# Show a few rows:
knitr::kable(head(mtcars_df))
```

Afterwards, always disconnect from the DBMS, stop the docker container and (optionally) remove it.
```{r}
dbDisconnect(con)
system2("docker", "stop temporary-postgres", stdout = TRUE, stderr = TRUE)
system2("docker", "rm temporary-postgres", stdout = TRUE, stderr = TRUE)
```

If we `stop` the docker container but don't remove it (with the `rm temporary-postgres` command), the container will persist and we can start it up with `start temporary-postgres`.  In that case, `mtcars` would still be there and we could download it again.  Since we have now removed the `temporary-postgres` container, the whole database has been deleted.  (There are enough copies of `mtcars` in the world, so no great loss.)
