# Docker, Postgres, and R

We always load the tidyverse and some other packages, but don't show it unless we are using packages other than `tidyverse`, `DBI`, `RPostgres`, and `glue`.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(DBI)
library(RPostgres)
library(glue)
```
## Verify that Docker running

Docker commands can be run from a terminal (e.g., the Rstudio Terminal pane) or with a `system()` command.  In this tutorial, we use `system2()` so that all the output that is created externally is shown.  Note that `system2` calls are divided into several parts:

1. The program that you are sending a command to.
2. The parameters or commands that are being sent
3. `stdout = TRUE, stderr = TRUE` are two parameters that are standard in this book, so that the comand's full output is shown in the book.

The `docker version` command returns the details about the docker daemon that is running on your computer.
```{r}
system2("docker", "version", stdout = TRUE, stderr = TRUE)
```

## Clean up if appropriate
Remove the `cattle` and `sql-pet` containers if they exists (e.g., from a prior experiments).  
```{r docker cleanup}
if (system2("docker", "ps -a", stdout = TRUE) %>% 
   grepl(x = ., pattern = 'cattle') %>% 
   any()) {
     system2("docker", "rm -f cattle")
}
if (system2("docker", "ps -a", stdout = TRUE) %>% 
   grepl(x = ., pattern = 'sql-pet') %>% 
   any()) {
     system2("docker", "rm -f sql-pet")
}
```

The convention we use in this book is to assemble a command with `glue` so that the you can see all of its separate parts.  The following chunk just constructs the command, but does not execute it.  If you have problems executing a commnad, you can always copy the command and execute in your terminal session.
```{r glue docker command, eval=TRUE, echo=TRUE}
docker_cmd <- glue(
  "run ",      # Run is the Docker command.  Everything that follows are `docker run` parameters.
  "--detach ", # (or `-d`) tells Docker to disconnect from the terminal / program issuing the command
  "--name cattle ",       # tells Docker to give the container a name: `cattle`
  "--publish 5432:5432 ", # tells Docker to expose the Postgres port 5432 to the local network with 5432
  " postgres:10"  # tells Docker the image that is to be run (after downloading if necessary)
)

# We name containers `cattle` for throw-aways and `pet` for ones we treasure and keep around.  :-)
```

Submit the command constructed above:
```{r run the docker command, eval=TRUE, echo=TRUE}
docker_cmd

system2("docker", docker_cmd, stdout = TRUE, stderr = TRUE)
```

Docker returns a long string of numbers.  If you are running this command for the first time, Docker is downloading the Postgres image and it takes a bit of time.

The following comand shows that `postgres:10` is still running:
```{r, eval=TRUE, echo=TRUE}
system2("docker", "ps", stdout = TRUE, stderr = TRUE)

```
## Connect, read and write to Postgres from R

Create a connection to Postgres waiting up to 10 seconds for the database to come up in Docker.

```{r}
wait_for_postgres <- function(seconds_to_test){
  for (i in 1:seconds_to_test) {
    db_ready <- DBI::dbCanConnect(RPostgres::Postgres(),
                        host = "localhost",
                        port = "5432",
                        user = "postgres",
                        password = "postgres")
    if ( !db_ready ) {Sys.sleep(1)} 
    else {con <- DBI::dbConnect(RPostgres::Postgres(),
                        host = "localhost",
                        port = "5432",
                        user = "postgres",
                        password = "postgres")
    }
    if (i == seconds_to_test & !db_ready) {con <- "it's not there"}
    }
    con
  }

con <- wait_for_postgres(10)

```

Show that you can connect but that Postgres database doesn't contain any tables:

```{r}
dbListTables(con)
```

Write `mtcars` to Postgres
```{r}
dbWriteTable(con, "mtcars", mtcars)
```

List the tables in the Postgres database to show that `mtcars` is now there:

```{r}
dbListTables(con)

# list the fields in mtcars:
dbListFields(con, "mtcars")
```

Download the table from the DBMS to a local data frame:
```{r}
mtcars_df <- tbl(con, "mtcars")

# Show a few rows:
knitr::kable(head(mtcars_df))
```

## Clean up

Afterwards, always disconnect from the DBMS, stop the docker container and (optionally) remove it.
```{r}
dbDisconnect(con)

# tell Docker to stop the container:
system2("docker", "stop cattle", stdout = TRUE, stderr = TRUE)

# tell Docker to remove the container from it's library of active containers:
system2("docker", "rm cattle", stdout = TRUE, stderr = TRUE)
```

If we `stop` the docker container but don't remove it (with the `rm cattle` command), the container will persist and we can start it up again later with `start cattle`.  In that case, `mtcars` would still be there and we could retrieve it from R again.  Since we have now removed the `cattle` container, the whole database has been deleted.  (There are enough copies of `mtcars` in the world, so no great loss.)
