# A persistent database in Postgres in Docker - piecemeal

## Overview

This chapter essentially repeats what was presented in the previous one, but does it in a step-by-step way that might be useful to understand how each of the steps involved in setting up a persistent Postgres database works.  If you are satisfied with the method shown in that chapter, skip this one for now.

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(DBI)
library(RPostgres)
```
## Retrieve the backup file

The first step is to get a local copy of the `dvdrental` Postgres restore file.  It comes in a zip format and needs to be un-zipped.  Use the `downloader` and `here` packages to keep track of things.
```{r}
if (!require(downloader)) install.packages("downloader")
if (!require(here)) install.packages("here")
library(downloader, here)

download("http://www.postgresqltutorial.com/wp-content/uploads/2017/10/dvdrental.zip", destfile = here("dvdrental.zip"))

unzip(here("dvdrental.zip"), exdir = here()) # creates a tar archhive named "dvdrental.tar"

file.remove(here("dvdrental.zip")) # the Zip file is no longer needed.

```

## Now, verify that Docker is up and running:
```{r docker verify}

system2("docker", "version", stdout = TRUE, stderr = TRUE)
```

Remove the `pet` container if it exists (e.g., from a prior run)
```{r}
if (system2("docker", "ps -a", stdout = TRUE) %>% 
   grepl(x = ., pattern = 'pet') %>% 
   any()) {
     system2("docker", "rm -f pet")
}
```

## Build the Docker Image

Build an image that derives from postgres:10.  Connect the local and Docker directories that need to be shared.  Expose the standard Postgres port 5432.
```{r}
wd <- getwd()

docker_cmd <- paste0(
  "run -d --name pet --publish 5432:5432 ",
  '--mount "type=bind,source=', wd,
  '/,target=/petdir"',
    " postgres:10"
)

system2("docker", docker_cmd, stdout = TRUE, stderr = TRUE)
```

Peek inside the docker container and list the files in the `petdir` directory.  Notice that `dvdrental.tar` is in both.
```{r}
system2('docker', 'exec pet ls petdir | grep "dvdrental.tar" ',
        stdout = TRUE, stderr = TRUE)

dir(wd, pattern = "dvdrental.tar")
```

We can execute programs inside the Docker container with the `exec` command.  In this case we tell Docker to execute the `psql` program inside the `pet` container and pass it some commands.
```{r}
Sys.sleep(2)
# inside Docker, execute the postgress SQL command-line program to create the dvdrental database:
system2('docker', 'exec pet psql -U postgres -c "CREATE DATABASE dvdrental;"',
        stdout = TRUE, stderr = TRUE)
```
The `psql` program repeats back to us what it has done, e.g., to create a databse named `dvdrental`.

Next we execute a different program in the Docker container, `pg_restore`, and tell it where the restore file is located.  If successful, the `pg_restore` just responds with a very laconic `character(0)`.
```{r}
Sys.sleep(2)
# restore the database from the .tar file
system2("docker", "exec pet pg_restore -U postgres -d dvdrental petdir/dvdrental.tar", stdout = TRUE, stderr = TRUE)

file.remove(here("dvdrental.tar")) # the tar file is no longer needed.
```

Use the DBI package to connect to Postgres.  But first, wait for Docker & Postgres to come up before connecting.
```{r}

Sys.sleep(4) 

con <- DBI::dbConnect(RPostgres::Postgres(),
                      host = "localhost",
                      port = "5432",
                      user = "postgres",
                      password = "postgres",
                      dbname = "dvdrental" ) # note that the dbname is specified

dbListTables(con)

dbListFields(con, "rental")
dbDisconnect(con)
```
## Stop and start to demonstrate persistence

Stop the container
```{r}
system2('docker', 'stop pet',
        stdout = TRUE, stderr = TRUE)
```
Restart the container and verify that the dvdrental tables are still there
```{r}

system2("docker",  "start pet", stdout = TRUE, stderr = TRUE)

Sys.sleep(1) # need to wait for Docker & Postgres to come up before connecting.

con <- DBI::dbConnect(RPostgres::Postgres(),
                      host = "localhost",
                      port = "5432",
                      user = "postgres",
                      password = "postgres",
                      dbname = "dvdrental" ) # note that the dbname is specified

glimpse(dbReadTable(con, "rental"))

```

Stop the container & show that the container is still there, so can be started again.
```{r}
system2('docker', 'stop pet',
        stdout = TRUE, stderr = TRUE)

# show that the container still exists even though it's not running
psout <- system2("docker", "ps -a", stdout = TRUE)
psout[grepl(x = psout, pattern = 'pet')]
```

Next time, you can just use this command to start the container:

`system2("docker",  "start pet", stdout = TRUE, stderr = TRUE)`

And after disconnecting from it the container can be completely removed with:

`system2("docker",  "rm pet -f", stdout = TRUE, stderr = TRUE)`
