# Postgres Examples, part B

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(DBI)
library(RPostgres)
library(glue)
library(knitr)
```

```{r echo=FALSE}
read_chunk('book-src/shared-db-login-code.R')

source('book-src/shared-db-login-code.R')  # use when debugging outside of knitr
```

```{r get_postgres_connection, eval=TRUE, echo=FALSE}

```

## Verify Docker is up and running:
```{r}
result <- system2("docker", "version", stdout = TRUE, stderr = TRUE)
result
```
verify pet DB is available, it may be stopped.
```{r}
result <- system2("docker", "ps -a", stdout = TRUE, stderr = TRUE)
result

any(grepl('Up .+pet$',result))

```
Start up the `docker-pet` container
```{r}
result <- system2("docker", "start sql-pet", stdout = TRUE, stderr = TRUE)
result


```


now connect to the database with R
```{r}
# need to wait for Docker & Postgres to come up before connecting.


#change the method here.
con <- wait_for_postgres(user = Sys.getenv("DEFAULT_POSTGRES_USER_NAME"),
                         password = Sys.getenv("DEFAULT_POSTGRES_PASSWORD"),
                         dbname = "dvdrental",
                         seconds_to_test = 10)


```

```{r}
## meta data: check existence of a table
rs1 <- dbGetQuery(con
                 ,"SELECT c.* 
                     FROM pg_catalog.pg_class c
                     JOIN pg_catalog.pg_namespace n ON n.oid = c.relnamespace
                    WHERE  n.nspname = 'public'
                      AND  c.relname = 'cust_movies'
                      AND  c.relkind = 'r'
                   ;
                 "
                 )
head(rs1)
```


```{r}
## create table via SQL statement
rs <- dbGetQuery(con
                ,'CREATE TABLE cust_movies AS
                    select c.customer_id
                          ,first_name
                          ,last_name,title
                          ,description
                      from customer c join rental r on c.customer_id = r.customer_id
                        join inventory i on r.inventory_id = i.inventory_id 
                        join film f on i.film_id = f.film_id 
                    order by last_name,first_name;
                 '
                )
head(rs)
```
Moved the following lines from #13
```{r}
## how many customers are there in the DVD Rental System
rs1 <- dbGetQuery(con,'select * from customer;')
kable(head(rs1))

pco <- dbSendQuery(con,'select * from customer;')
rs2  <- dbFetch(pco)
dbClearResult(pco)
kable(head(rs2))

```


```{r}
pco <- dbSendQuery(con,"select * from customer where customer_id between $1 and $2")
dbBind(pco,list(501,525))
rs2  <- dbFetch(pco)
dbClearResult(pco)
kable(head(rs2))
```

```{r}
rs1 <- dbGetQuery(con,
                "explain select r.*
                   from rental r 
                 ;"
                )  
head(rs1)

rs2 <- dbGetQuery(con,
                "explain select count(*) count
                   from rental r 
                        left outer join payment p 
                          on r.rental_id = p.rental_id  
                    where p.rental_id is null
                 ;")
head(rs2)

rs3 <- dbGetQuery(con,
                "explain select sum(f.rental_rate) open_amt,count(*) count
                   from rental r 
                        left outer join payment p 
                          on r.rental_id = p.rental_id 
                        join inventory i
                          on r.inventory_id = i.inventory_id
                        join film f
                          on i.film_id = f.film_id
                    where p.rental_id is null
                 ;")
head(rs3)

rs4 <- dbGetQuery(con,
                "explain select c.customer_id,c.first_name,c.last_name,sum(f.rental_rate) open_amt,count(*) count
                   from rental r 
                        left outer join payment p 
                          on r.rental_id = p.rental_id  
                        join inventory i
                          on r.inventory_id = i.inventory_id
                        join film f
                          on i.film_id = f.film_id
                        join customer c
                          on r.customer_id = c.customer_id
                  where p.rental_id is null
                  group by c.customer_id,c.first_name,c.last_name
                  order by open_amt desc
                 ;"
                )  
head(rs4)
```

## SQL Execution Steps

*  Parse the incoming SQL query
*  Compile the SQL query
*  Plan/optimize the data acquisition path
*  Execute the optimized query / acquire and return data

```{r}
dbWriteTable(con, "mtcars", mtcars)  
rs <- dbSendQuery(con, "SELECT * FROM mtcars WHERE cyl = 4")
dbFetch(rs)
dbClearResult(rs)
```

```{r}
#Pass one set of values with the param argument:
rs <- dbSendQuery(con,"SELECT * FROM mtcars WHERE cyl = 4")
dbFetch(rs)
dbClearResult(rs)

# Pass multiple sets of values with dbBind():
rs <- dbSendQuery(con, "SELECT * FROM mtcars WHERE cyl = $1")
dbBind(rs, list(6L))
dbFetch(rs)
dbBind(rs, list(8L))
dbFetch(rs)
dbClearResult(rs)
```

```{r}
dbWriteTable(con, "cars", head(mtcars, 3))
dbReadTable(con, "cars")   # there are 3 rows

# not working at the moment.  Not sure what Sophie had in mind...  - JDS
# dbExecute(
#   con,
#   "INSERT INTO cars (speed, dist) VALUES ($1, $2)",
#   param = list(4:7, 5:8)
# )
dbReadTable(con, "cars")   # there are now 10 rows
```
 
Clean up
```{r}
dbRemoveTable(con, "cars")
dbRemoveTable(con, "mtcars")
dbRemoveTable(con, "cust_movies")

# diconnect from the db
dbDisconnect(con)

result <- system2("docker", "stop sql-pet", stdout = TRUE, stderr = TRUE)
result

```

